# Major acts and regulations

## Acts and laws of Nepal

\begin{solution}
\begin{enumerate}
\item Nepali Language Publication Corporation Act, 2021 (1964)
\item Nepal Ship (Registration and Logbook) Act, 2027 (1971)
\item Nepal Special Service Act, 2042 (1985)
\item Non-resident Nepali Act, 2064
\item Passports Act, 2024 (1967)
\item The Consumer Protection Act, 2075 (2018)
\item Work Operation Fund Act, 2043 (1986)
\item Water Tax Act, 2023 (1966)
\item Nepal Treaty Act, 2047 (1990)
\item Supreme Court Act, 2048 (1991)
\item Regional Development Plans (Implementation) Act,2013 (1956)
\item Rajya Rajauta (Petty States and Rulers) Act, 2017 (1961)
\item Periodic Tax Recovery Act, 2012 (1955)
\item The University Grants Commission Act, 2050 (1993 A.D.)
\item The Stamp Duty Act, 2019 (1963)
\item The Pesticides Act, 2048 (1991)
\item Radio Act, 2014 (1957)
\item Act Restricting Investment Abroad, 2021(1964)
\item Act to Provide for Enhancement of the Circulation of Nepalese Currency, 2014(1957)
\item The Administrative Procedures (Regulation) Act, 2013 (1956)
\item Animal Health and Livestock Services Act, 2055 (1999)
\item Animal Slaughterhouse and Meat Inspection Act, 2055(1999)
\item Anti Defection Act, 2054 (1997)
\item Contingency Fund Act, 2016 (1959)
\item Drugs Act, 2035, (1978)
\item Nepal Tourism Board Act, 2053
\item Public Security Act, 2046 (1989)
\item Associations Registration Act, 2034 (1977)
\item AUDIT ACT, 2048 (1991)
\item banking-offence-and-punishment-act-2064-2008
\item Child Labor (Prohibition and Regulation) Act, 2056 (2000)
\item Commissions of Inquiry Act, 2026 (1969)
\item Infectious Disease Act, 2020(1964)
\item Immovable Property Requisition Act, 2013 (1956)
\item Human Trafficking and Transportation (Control) Act, 2064
\item The Copyright Act, 2059 (2002)
\item Development Board Act, 2013 (1956)
\item Domestic Postal Goods (Insurance) Act, 2019 (1962)
\item Foreign Investment Tax Act, 2019 (1963)
\item Judicial Fund Act, 2043 (1986)
\item Nepal Citizenship Act, 2063 (2006)
\item Essential Services Operation Act, 2014 (1957)
\item Export and Import (Control) Act, 2013 (1957)
\item Essential Goods Protection Act, 2012 (1955)
\item Extradition Act, 2045 (1988)
\item Legal Aid Act, 2054 (1997)
\item Nepal Agency Act 2014
\item International Financial Transactions Act, 2054 (1998)
\item National Judicial Academy Act, 2063 (2006)
\item Mother's Milk Substitutes (Control of Sale and Distribution) Act, 2049 (1992)
\item Compensation Relating to Torture Act, 2053 (1996)
\item Multimodal Transportation of Goods Act, 2063(2006).
\item National Trust for Nature Conservation Act, 2039 (1982)
\item Electricity Act, 2049 (1992)
\item Interpretation of Laws Act, 2010(1954)
\item Lottery Act, 2025 (1968)
\item Nepal Electricity Authority Act, 2041 (1984)
\item Constitutional Council (Functions, Duties, Powers and Procedures) Act, 2066 (2010)
\item Feed Act, 2033 (1976)
\item Jhora Area Land Act, 2028 (1971)
\item National Parks and Wildlife Conservation Act, 2029 (1973)
\item National Guidance Act, 2018 (1961)
\item Loan and Guarantee Act, 2025(1968)
\item Narcotic Drugs (Control) Act, 2033 (1976)
\item Motor Vehicle Tax Act, 2031 (1974)
\item Nepal Health Research Council Act, 2047 (1991)
\item Competition Promotion and Market Protection Act, 2063 (2007)
\item Evidence Act, 2031
\item Money Laundering Prevention Act, 2063 (2008)
\item Motion Picture (Production, Exhibition and Distribution) Act, 2026 (1969)
\item Arbitration Act, 2055 (1999)
\item Domestic Violence (Offence and Punishment) Act, 2066 (2009)
\item Donations Act, 2030 (1973)
\item National Cooperatives Development Board Act, 2049 (1992)
\item Insurance Act, 2049 (1992)
\item Aquatic Animal Protection Act, 2017(1960)
\item National Dairy Development Board Act, 2048(1992)
\item Explosives Act, 2018 (1961)
\item Essential Commodities Control (Authorization) Act, 2017 (1961)
\item Impeachment (Regulation of Procedure) Act, 2059 (2002)
\item Birth, Death and Other Personal Events (Registration) Act, 2033 (1976)
\item The Foreign Investment and Technology Transfer Act, 2019 (2075)
\item House and Land Tax Act, 2019 (1962)
\item The Industrial Enterprises Act, 2049 (1992)
\item Nepal Standards (Certification Mark) Act, 2037 (1980)
\item National Tea and Coffee Development Board Act, 2049 (1993)
\item Bonded Labour ( Prohibition )Act, 2058 (2002)
\item Diplomatic Privilege and Immunities of the Foreign States and Representatives Act, 2027 (1970)
\item The National Broadcasting Act, 1993
\item Industrial Enterprises Development Institute Act, 2053 (1996)
\item Lumbini Development Trust Act, 2042 (1985)
\item Act Prohibiting the Charging and Taking of Amount (Bhatti Charsa etc.) by Birta-holder on Birta, 2015 (1958)
\item Act Relating to Institutions Acting as Financial Intermediary, 2055 (1999)
\item Act Relating to Land of Jhora Area, 2028 (1971)
\item Nepal Agricultural Research Council Act, 2048(1992)
\item Agriculture and Forestry University Act, 2067 (2010)
\item Nepal Water Supply Corporation Act, 2046 (1989)
\item Nepal Trust Act, 2064 (2008)
\item Nepal Pharmacy Council Act, 2057 (2000)
\item Nepal Nursing Council Act, 2052 (1996)
\item Nepal Rastra Bank Act, 2058 (2002)
\item Nepal Medical Council Act, 2020 (1964)
\item Nepal Health Professional Council Act, 2053
\item Nepal Bar Council Act, 2050 (1993)
\item Nepal Law Commission Act, 2063 (2007 )
\item Nepal Airlines Corporation Act, 2019 (1963)
\item National Human Rights Commission Act, 2068 (2012)
\item National Insurance Corporation Act, 2025 (1968)
\item Nepal Administrative Staff College Act, 2020 (1982)
\item National News Agency Act, 2019 (1962)
\item National Foundation for Upliftment of Aadibasi/Janjati Act, 2058 (2002)
\item Nepal Music and Dance Academy Act, 2064 (2007)
\item Mines and Minerals Act, 2042 (1985)
\item Mediation Act, 2068 (2011)
\item Archives Preservation Act, 2046 (1989)
\item International Finance Corporations Membership Acquisition Act, 2022 (1965)
\item Corporations Act, 2021 (1964)
\item Black Marketing and Some Other Social Offenses and Punishment Act, 2032 (1975)
\item Caste Based Discrimination and Untouchability Offence and Punishment Act, 2068 (2011)
\item Forest Act, 2049 (1993)
\item Construction Business Act, 2055 (1999)
\item Bretton Woods Agreement Act, 2018 (1961)
\item Far-western University Act, 2067 (2010)
\item Negotiable Instruments Act, 2034 (1977)
\item Mid-western University Act, 2067 (2010)
\item Gorkhapatra Corporation Act, 2019 (1963)
\item Hotel Management and Sale and Distribution of Liquors (Control) Act, 2023 (1966)
\item The B. P. Koirala Institute of Health Sciences Act, 2049 (1993)
\item The Greater Janakpur Area Development Council Act, 2055 (1998)
\item Iodized Salt (Production, Sale and Distribution) Act, 2055(1998)
\item Amount and Governmental Contract Arrangement Act, 2020(1963)
\item Insolvency Act, 2063 (2006)
\item Land Acquisition Act, 2034 (1977)
\item Kathmandu University Act, 2048 (1991)
\item International Development Association's Membership Acquisition Act, 2019 (1962)
\item Land Revenue Act, 2034 (1978)
\item Martyr Gangalal National Heart Disease Centre Act, 2057 (2000)
\item Lumbini Bouddha University Act, 2067 (2006)
\item Land (Survey and Measurement) Act, 2019 (1963)
\item Local Administration Act, 2028 (1971)
\item Local Development Training Academy Act, 2049 (1993)
\item International Centre for Integrated Mountain Development Act, 2040 (1983)
\item The Human Body Organ Transplantation (Regulation and Prohibition) Act, 2055 (1998)
\item The Guthi Corporation Act, 2033 (1976)
\item Ancient Monument Preservation Act, 2013 (1956)
\item The Electronic Transactions Act, 2063 (2008)
\item Armed Police Force Act, 2058 (2001)
\item Arms and Ammunition Act, 2019 (1962)
\item Ayurveda Medical Council Act, 2045 (1988)
\item The B.P. Koirala Memorial Cancer Hospital Act, 2053(1997)
\item Birta Abolition Act, 2016 (1959)
\item The Building Act, 2055
\item Nepal Chartered Accountants Act, 2053 (1997)
\item Citizen Investment Trust Act, 2047 (1991)
\item Commission for the Investigation of Abuse of Authority Act, 2048 (1991)
\item Communication Corporation Act, 2028 (1972)
\item Compensation Act, 2019 (1963)
\item The Lands Act, 2021(1964)
\item Construction Business Act, 2055 (1999)
\item Corporations Act, 2021 (1964)
\item Customs Act, 2064 (2007)
\item Decorations Act, 2064 (2007)
\item Education Act, 2028 (1971)
\item Electricity Theft Control Act, 2058 (2002)
\item Excise Duty Act, 2058 (2002)
\item Nepal Fine Arts Academy Act, 2064 (2007)
\item Privatization Act, 2050 (1994)
\item Water Resources Act, 2049 (1992)
\item Private Forests Nationalization Act, 2013 (1957)
\item The Social Welfare Act, 2049 (1992)
\item Notary Public Act, 2063
\item Right to Information Act, 2064 (2007 )
\item Trade Union Act, 2049 (1992)
\item Tax Settlement Commission Act, 2033 (1976)
\item Telecommunications Act, 2053 (1997)
\item Statistics Act, 2015 (1958)
\item Working Journalists Act, 2051(1993)
\item Wealth Tax Act, 2047 (1990)
\item Water Tariff Fixation Commission Act, 2063 (2006)
\item Water Supply Management Board Act, 2063 (2006)
\item Nepal Veterinary Council Act, 2055 (1999)
\item Ukhada Act, 2021 (1964)
\item Tribhuvan University Act, 2049(1992)
\item Town Development Fund Act, 2053 (1997)
\item Town Development Act, 2045 (1998)
\item Road Board Act, 2058 (2002)
\item Postal Act, 2019 (1963)
\item Partnership Act, 2020 (1964
\item Police Act, 2012 (1955)
\item The Patent, Design and Trade Mark Act, 2022 (1965)
\item Special Court Act, 2059 (2002)
\item [The] Kathmandu Valley Development Authority Act, 2045 B.S. (1988)
\item Technical Education and Vocational Training Council Act, 2045 (1988)
\item Summary Procedures Act, 2028 (1972)
\item Standard Measurement and Weight Act, 2025 (1968)
\item Sports Development Act, 2048 (1992)
\item Some Public Documents Authentication (Procedures) Act, 2063 (2006)
\item Solid Waste Management Act, 2068 (2011)
\item Soil and Watershed Conservation Act, 2039(1982)
\item Social Practices (Reform) Act, 2033 (1976)
\item Ship Registration Act, 2027 (1971)
\item Senior Citizens Act, 2063 (2006)
\item Seeds Act, 2045 (1988)
\item Tourism Act, 2035 (1978)
\item Tobacco Products (Control and Regulatory) Act, 2068 (2011)
\item [The] Ownership of Joint Housing Act, 2054 B.S. (1997)
\item Security of the Health Workers and Health Organizations Act, 2066 (2010)
\item Secured Transaction Act, 2063 (2006)
\item Scholarships Act, 2021 (1964)
\item Revenue Tribunal Act, 2031 (1974)
\item Securities Act, 2063 (2007)
\item Press Council Act 2048 (1992) 
\item Patan Academy of Health Sciences Act, 2064 (2008)
\item The Prevention of Corruption Act, 2059 (2002 A.D)
\item Purwanchal University Act, 2050 (1994)
\item Public Service Commission Act, 2066 (2010)
\item Public Roads Act, 2031 (1974)
\item The Public Procurement Act, 2063 (2007)
\item Public Debt Act, 2059 (2002)
\item Prisons Act, 2019 (1963)
\item Remuneration, Conditions of Service and Facilities of the Authorities of the Constitutional Bodies Act, 2053 (1997)
\item Press and Publication Act, 2048 (1991)
\item Poverty Alleviation Fund Act, 2063 (2006)
\item Pokhara University Act, 2053 (1996)
\item Plant Protection Act, 2064 (2007)
\item Nepal Petroleum Act, 2040 (1983)
\item Pasture Land Nationalization Act, 2031 (1974)
\item Mutual Legal Assistance Act, 2070 (2014)
\item Judicial Council Act, 2073 (2016)
\item Employee Provident Fund Act, 2019 (1962)
\item An Act to Regulate and Control International Trade in Endangered Wild Fauna and Flora
\item Administration of Justice Act, 2073 (2016)
\item Pashupati Area Development Trust Act, 2044 (1987)
\item Karnali Academy of Health Sciences Act, 2068 (2011)
\item Nepal Health Service Act, 2053 (1997)
\item Army Act, 2063 (2006 )
\item Civil Service Act, 2049 (1993)
\item Motor Vehicles and Transport Management Act, 2049 (1993)
\item Passports Act, 2024 (1967)
\item The Crime Victim Protection Act, 2075 (2018)
\item The Privacy Act, 2075 (2018)
\item The Public Health Service Act, 2075 (2018)
\item The Right to Employment Act, 2075 (2018)
\item The Act Relating to Rights of Persons with Disabilities, 2074 (2017)
\item The Right to Food and Food Sovereignty Act, 2075 (2018)
\item The Right to Housing Act, 2075 (2018)
\item The Right to Safe Motherhood and Reproductive Health Act, 2075 (2018)
\item The Social Security Act, 2075 (2018)
\item The Act Relating to Children, 2075 (2018)
\item The Act Relating to Compulsory and Free Education, 2075 (2018)
\item Bank and Financial Institution Act, 2073 (2017)
\end{enumerate}
\end{solution}

## Right to information

\begin{questions}

\question What are the provisions in Nepal relating on the "Right to Information" ? Discuss in detail.

\begin{solution}
Government of Nepal enacted the Right to Information Act, 2064 (2007)

\textbf{Preamble}

1. To make the functions of the state open and transparent in accordance with the democratic system and to make responsible and accountable to the citizen
2. To make the access of citizens simple and easy to the information of public importance held in public bodies
3. To protect sensitive information that could make adverse impact on the interest of the nation and citizen, and for the necessity to have legal provisions to protect the right of the citizen to be well-informed and to bring it into practice.

\textbf{Right to information}

1. Every citizen shall have the right to information subject to RTI act.
2. Every citizen shall have access to the information held in the public bodies.

\textbf{Provision regarding commission}

1. An independent national information commission will be established for the protection, promotion and practice of right to information.
2. A Chief Information Commissioner and two other information commissioners will be in the commission.
3. In order to appoint Chief Information Commissioner and Information Commissioners, there will be a committee comprised as follows:
  a. The speaker -- Chairman
  b. Minister of state minister for information and commission -- Member
  c. President, federation of Nepalese journalists -- Member
4. The Government of Nepal on the recommendation of the Committee under Sub-Section (3), will appoint a Chief Information Commissioner and Information Commissioners; while appointing in this way, at least one female will have to be included.
5. The Committee in accordance with Sub-Section (3) will follow inclusive principles as much as possible with recommending for appointment of Chief Information Commissioner and Information Commissioners.
6. The procedures of recommendation in accordance with Sub-Section (3) will be as determined by the Recommendation Committee itself.

Why RTI

RTI Act 2007 gives you the right to ask for and receive information about any activity from a public body. Public bodies include the following entities:

1. Bodies formed under the constitution
2. Bodies established by the Act
3. Bodies constituted by the government of Nepal
4. Public service provider organizations or foundations established as per the law
5. Political party and organization registered under prevalent laws
6. Organized institutions under the full or partial ownership or under control of the GoN, or organized bodies receiving grants from the GoN
7. Organized institutions formed by the Bodies established by the GoN or any other agencies lawfully formed through an agreement
8. NGOs/Institutions directly or indirectly funded by the GoN or foreign government or international organizations/institutions
9. Other bodies or institutions defined as public bodies by the GoN through a notice published in Nepal gazette.

The seeker of information has to submit a written application to the concerned information officer. It is important for the applicant to clearly specify the information sought. Unclear requests make things difficult to provide correct information. Hence, requests for information should be as clear and specific as possible.

No fees will be charged for the first half hour of monitoring in public bodies or getting information from work place, getting records, or undertaking records inspection and observation. After that the charge in Rs 50 per person, per hour. Likewise, there are no fees for inspection in public libraries and public places where information is made available.

Public bodies will nominate a staff as an information officer for the public's ease of access to information. The head of the office should provide information to the PIO to disseminate it in accordance to sub-section (I) of RTI act. An information section will be set up in all public bodies to disseminate information. The PIOs regularly update their information and keeps records regarding their work plans, expenditure, financial transactions, rules, regulations, laws, and directives. The RTI act anticipates that public bodies store (archive) their information going as far back as 20 years from the commencement of the RTI act. Unless there are resonable and sufficient grounds to validate those exceptions, public bodies cannot evade the responsibility to disclose information. Following information, however, shall not be obliged to be made available:

1. Information harmful to soverignty, integrity, national security public security, law and order of the country; and information that can jeopardize international relations;

2. Information that could have direct bearing on research, investigation and prosecution of a criminal case

3. Information harmful to economic, financial, monetary wellbeing or posing threat to someone's intellectual property rights, as also to banking and financial secrecy.

4. Information that could jeopardize the harmonious realations subsisting between various castes and communities.

5. Information posing threat to a person's right to privacy, personal security, health and property.


In case decision rendered by the Chief of Public body is not satisfactory, one can file the appeal (within 35 days of decision received) to the commission either by directly approaching the commission office or getting it registered in the concerned office or through the concerned District Administration Office. Alternatively, one can also mail the appeal directly to NIC, via any postal services.

Centre for Law and Democracy, Canada has published a comparative study report on existing RTI laws of 111 countries in the world, in which Nepal has been ranked 27th in the world with the score of 104 out of 150. The worldwide experience shows that RTI supports good governance and helps to reduce corruption. The National Information Commission had completed its first RTI Audit in 2016 through ex-secretaries of the government. The second RTI Audit was completed in 2017 through senior journalists.

After analysing all aspects, it was found that 11 ministries (35\%) were found working as per law. Similarly, 20 ministries (65\%) were partially satisfactory. The commission has made 10 suggestions, including adding "imparting information" in Article 27 of the constitution, taking out of the word "secret" from the employees' oath, amongst other, to the government on the basis of the results of this RTI audit.

\end{solution}

\end{questions}

## Army Act, 2063

\begin{questions}

\question According to Army Act, 2063 what are the punishment provisioned through Court Marshall ?

\begin{solution}
According to the Army Act, 2063 (Chapter-11; Section (dafa) 101), 

1. Following penalties may be imposed by court marshall:

\begin{itemize}
\item Life imprisonment along with confiscation of entire property,
\item Confiscation of his or her entire share of ancestral property,
\item Imprisonment of upto Fourteen years,
\item Removing from service without prejudicing eligibility for any governmental service in the future, or dismissing with ineligibility for any governmental service in the future,
\item Demotion to a lower rank,
\item Freezing of promotion for Five years,
\item Deduction of salary and allowances for Three months in case the offence has been committed in the course of military action,
\item Freezing of salary and allowances until the recovery of compensation,
\item Freezing of salary increments for a period not exceeding Two years,
\item Warning (Nasihat).
\end{itemize}

2. The following penalties shall be imposed for the following offences:

\begin{enumerate}
  \item Life imprisonment along with confiscation of entire property for the commission of offences as referred to in Section 38, 39 or Clause (a) of Sub-section (1) of Section 42.
  \item A maximum Fourteen years of imprisonment shall be imposed for the commission of the following offences:
  \begin{enumerate}
    \item Offences provided in Section 40,
    \item Offences provided in Section 41 and Clause (a) of Section 45 during a military operations,
    \item Offences provided in Section 43 or 46 if committed intentionally,
    \item Offences provided in Clause (a) or (b) of Section 44 committed against a person while discharging his/her duties during a military operation,
    \item Offences provided in Section 47.
  \end{enumerate}
  \item A maximum Ten years of imprisonment shall be imposed for the commission of the following offences:
  \begin{enumerate}
    \item Offences provided in Clauses (a) or (b) of Section 44 committed against a person not during the military operation,
    \item Offences provided in Section 48, 49 or 50, (3) Offences provided in Clause (a) of Section 51 or Clause (a) of Section 52.
  \end{enumerate}
  \item The penalty of confiscation of entire property or maximum of Seven years of imprisonment shall be imposed for the commission of offences provided in Clause (b) of Sub-section (1) of Section 42.
  \item A maximum of Seven years of imprisonment shall be imposed for the commission of the following offences:
  \begin{enumerate}
    \item Offences provided in Section 41 committed not during military operation,
    \item Offences provided in Section 46 that are committed unintentionally,
    \item Offences provided in Clauses (b), (c), (d), (e), (f) or (g) of Section 52,
    \item Offences provided in Section 43, 54 or 63.
  \end{enumerate}
  \item A maximum of Five years of imprisonment shall be imposed for the commission of the following offences:
  \begin{enumerate}
    \item Offences provided in Clause (c) of Section 44, (2) Offences provided in Clause (a) of Section 45 committed not during the period of a military operation,
    \item Offences provided in Clauses (b) or (c) of Section 51 or Sections 45, 56 or 57.
  \end{enumerate}
  \item A maximum Three years of imprisonment shall be imposed for the commission of offences provided in Clauses (c) or (d) of Sub-section (1) of Section 42 or Sections 58, 59 or 60.
  \item A maximum of Two years of imprisonment shall be imposed for the commission of the following offences:
  \begin{enumerate}
    \item Offences provided in Clauses (e), (f), (g) or (h) of Sub-section (1) of Section 42,
    \item Offences provided in Clause (b) of Section 45,
    \item Offences provided in Clause (h) of Section 52 committed by personnel below the rank of officer while discharging duties during a military operation,
    \item Offences provided in Section 43 that are committed unintentionally,
    \item Offences provided in Section 61.
  \end{enumerate}
  \item A maximum of Six months imprisonment shall be imposed if a personnel below the rank of officer commits offences provided in Clause (h) of Section 52 except in the period of military operation or while on duty.
  \item A person who commits offences provided in Clauses (i) and (j) of Section 52 shall be removed from the service.
  \item A person who commits offences provided in Clause (k) of Section 62 shall be barred from Two salary increments or promotion for Two years.
  \item A person who attempts to commit such offences as referred to in Chapter-7 or induces another person to commit such offences shall, unless there is specific provision in this Act, be imposed the following penalties:
  \begin{enumerate}
    \item A maximum of Fourteen years of imprisonment for attempting to commit an offence for which a penalty of life imprisonment along with confiscation of entire property may be imposed, and half of the maximum (range) imprisonment for other offences which have been attempted,
    \item Equal penalty for inducement (incitement) to commit an offence as imposed for committing an offence,
    \item A maximum of Fourteen years of imprisonment for inducement to commit an offence which has not been committed (completed), and half of the maximum (range) punishment for inducement to commit any other offence which had not actually been committed (completed).
    \item If a person who committed an offence provided in Clause (b) of Sub-section (1) of Section 42 has absconded and is not found (traced) does not produce him/herself within Thirty Five days, his/her partition share to ancestral property shall be frozen from being transacted. If such person does not appear him/herself or could be found through making a search within Three years, his/her partition share to ancestral property shall be confiscated and auctioned, and the money therefrom shall be deposited to the Government Treasury.
    \item After confiscation of partition share to ancestral property, if the  offender is found, he/she may be imposed life imprisonment along with confiscation of the entire property.
    \item A person who is dismissed from service and disqualified for any governmental job in future pursuant to Clause (j) of Sub-section (2), shall not be paid his/her remaining salary or allowances and pension which he/she is entitled and other facilities except the money accumulated in his/her provident fund.
  \end{enumerate}
\end{enumerate}

Besides these there is Battlefield penalty (Section 104) and Departmental action (Section 105) and Collective fine (Section 107).

\end{solution}

\end{questions}

## Forest Act, 2049

\begin{questions}

\question What provisions and guidelines have Forest act, 2049 set for utilization and management of forest in Nepal? Mention.

\begin{solution}

The Forest Act, 1992, defines national forest, state-managed forest, protected forest, community forest, leasehold forest, religious forest and private forest as well as ownership of different types of forest, and their management. The Act encourages people to own and to use forest, rather than depend on national or state-owned forest. The Forest Regulation, 1994, fixes users' price rates for timber, fuel-wood, medicinal plants and herbs, flowers and furs, fruits and seeds, bark, logs, and allied forest products. Likewise, charges for animal and cattle grazingare also fixed by the regulation.

Similarly, separate regulations for national parks such as the Royal Chitwan National Park and Royal Bardia National Park make it possible to charge different fees for entrance and recreational activities such as safari, hunting game species and fishing. Another milestone in the forestry sector is the Buffer Zone Management Regulation, 1995, which aims to harmonize relation between national parks and the people around those parks. This regulation also encourages the communities around parks to form users' groups so that they can mange and use the buffer zone resources.

\end{solution}

\end{questions}

## The National Civil (Code) Act, 2017 (2074)

Whereas, it is expedient to make timely the civil provisions contained in the country code and other laws, by also amending and consolidating such provisions, in order to maintain morality, decency, etiquette and convenience as well as economic interest of the public by maintaining law and order in the country and maintain harmonious relationship between various castes, races and communities, by making just provisions in the economic, social and cultural field;

Now, therefore, the Legislature-Parliament under clause (1) of Article 296 of the consitution of Nepal has enacted this act. 

- There are a total of 721 sections.

Part-1 Preliminary

Chapter-1 Preliminary statements

Chapter-2 General principles of civil law

- Principles and provisions applicable generally
- Ignorance of law not to be excused
- Act against public interest not to be carried out
- Act contrary to law to be invalid
- A wrongdoer to bear liability for loss or damage
- Act causing nuisance to other not to be done
- Not to enjoy benefit of wrong
- Act against interest to be invalid
- Personality to be recognized
- Good neighborhood to be maintained
- Not to be relieved from civil liability
- Custom or tradition contrary to law not to be recognized
- Applicable to foreigners as well

Chapter-3 Provisions relating to civil rights

- To be equal before law
- Discrimination not to be made
- Discrimination not to be deemed where special provision is made
- Guarantee of freedoms and rights
- Right to privacy deemed to be violated
- Right to make contract
- No taxation except in accordance with law
- No to employ in work against will
- Property not to be acquired, requisitioned, auctioned or forfeited
- Complaint may be filed
- Compensation to be awarded
- Compensation for filing false complaint
- Statute of limitation

Part-2 Law relating to persons

Chapter-1 Provisions relating to natural persons

Chapter-2 Provisions relating to legal persons

Chapter-3 Provisions relating to bankruptcy of natural persons

Part-3 Family law

Chapter-1 Provisions relating to marriage

Chapter-2 Provisions relating to consequence of marriage

Chapter-3 Provisions relating to divorce

Chapter-4 Provisions relating to relationship of parents and children

Chapter-5 Provisions relating to maternal and paternal authority

Chapter-6 Provisions relating to guardianship

Chapter-7 Provisions relating to curatorship

Chapter-8 Provisions relating to adoption

Chapter-9 Provisions relating to inter-country adoption

Chapter-10 Provisions relating to partition

Chapter-11 Provisions relating to succession

Part-4 Law relating to property

Chapter-1 General provisions relating to property

Chapter-2 Provisions relating to ownership and possession

Chapter-3 Provisions relating to use of property

Chapter-4 Provisions relating to cultivation, possession and registration of land

Chapter-5 Provision relating to government, public and community property

Chapter-6 Provision relating to trusts

Chapter-7 Provision relating usfruct

Chapter-8 Provisions relating to servitudes

Chapter-9 Provisions relating to house rent

Chapter-10 Provisions relating to gift and donation

Chapter-11 Provisions relating to transfer and acquisition of property

Chapter-12 Provisions relating to mortgage of immovable property

Chapter-13 Provisions relating to registration of deeds

Chapter-14 Provisions relating to transactions

Part-5 Provision relating to contracts and liabilities

Chapter-1 Provisions relating to obligations

Chapter-2 Provisions relating to formation of contracts

Chapter-3 Validity of contracts

Chapter-4 Provision relating to performance of contracts

Chapter-5 Provision relating to breach of contract and remedies

Chapter-6 Provision relating to contracts of sales of goods

Chapter-7 Provisions relating to contracts of guarantee

Chapter-8 Provisions relating to contracts of bailment

Chapter-9 Provisions relating to contracts of pledge or deposit

Chapter-10 Provisions relating to contracts of agency

Chapter-11 Provisions relating to contracts of carriage of goods

Chapter-12 Provisions relating to contracts of lease

Chapter-13 Provisions relating to Hire-purchase contracts

Chapter-14 Provisions relating to wages

Chapter-15 Provisions relating to indirect or quasi contracts

Chapter-16 Provisions relating to unjust enrichment

Chapter-17 Provisions relating to torts

Chapter-18 Provisions relating to liability for defective products

Part-6 Provisions relating to private international law

## Civil service act, 2049

Date of authentification and publication: 2050-02-04 (May 17, 1993)

1. Nepal health services act, 2053 (1997)
2. Civil service act (First amendment), 2055 (1998)
3. Some Nepal acts amendment act, 2057 (2000)
4. Parliament secreteriat act, 2058 (2001)
5. Civil service act (Second amendment), 2064 (2007)
6. Federal reinforcement and some national act amending acts, 2066
7. Civil service act (Third amendment), 2070
8. Civil service act (Forth amendment), 2072

Chapter-1 Preliminary

Chapter-2 Constitution of civil services

1. Relating to central employee administration agency
2. Organization of civil services
  - Nepal engineering service
  - Nepal agricultural service
  - (Nepal judicial service)
  - Nepal foreign service
  - Nepal audit service
  - Nepal forestry service
  - Nepal miscellenaous service
  - Nepal education service

3. Classes of civil service
  - Gazetted: Special, first, second, third
  - Non-gazetted: First, second, third, forth, fifth

Chapter-3 Fulfilment of vacancy of civil service

Chapter-4 Transfer and promotion

Chapter-5 Salary, allowance, dashain expenses and other facilities

Chapter-6 Retirement, gratuity and pension

- Any civil employee who has completed the age of fifty-eight years or completed the tenure shall ipso facto retire from the civil service.
- 

Chapter-7 Conduct

- Timeliness/punctuality and regularity (Timeliness specifications is included in Section 41 of Civil Service Rules, 2050; Condition of unpuntual arrival of an employee to be treated as absentee is included in Section 55)
- Discipline and obedience
- Political impact should not be made
- Non participation in politics
- Government should not be protested against
- Publication/reporting on government activities is prohibited
- Should not accept grant, present and donations
- Cannot establish company and/or business enterprise
- Cannot participate in election
- Cannot parttake in other employment or services
- Cannot seek permit for permanent residence
- Cannot parttake in demonstration and strikes
- Cannot partake in any activities that refrain from government carrying out activities
- Cannot represent trade union
- Provisions relating to organization of a trade union
- Should disclose his/her property details
- Cannot engage in torturing and extortion
- Other conduct
- Conduct towards public
- May be issued warning

Chapter-8 Service security

Chapter-9 Punishment and appeal

(Section 59) subjected to the department punishment, if there exist appropriate and sufficient reason.

**Ordinary punishment**

- Censure
- Withholding of promotion for up to two yeas or withholding of a maximum of two salary increments,
- Withholding of promotion for two to five years, or withholding of a maximum of two to five salary increments,

**Special punishment**

- Removal from service, without being disqualified for government service in the future.
- Dismissal from service, with being disqualified for government service in the future.

Chapter-10 Miscellaneous

(Section 71 and 71 A) Provisions for leave

The civil employees shall be entitled to the following leaves, as prescribed:

1. Casual and festival leave,
2. Home leave,
3. Sick leave,
4. Maternity leave,
5. Maternity care leave,
6. Obsequies leave,
7. Study leave,
8. Extra-ordinary leave

- If the husband or wife of any civil employee is deputed to the foreign based Nepalese Embassy or mission, such civil employee shall get unpaid leave.

(for more detailed discussion of provisions relating to leave, refer to Subsection 'Provisions for leave'.)

### Terms and definition of terms

\begin{questions}
\question What is family as per the Civil Services Act, 2049 ?

\begin{solution}

"Family" means the husband, wife, son, unmarried daughter, adpoted son, unmarried adopted daughter, father, mother or step mother living together with a Civil employee and whom he/she himself/herself has to maintain and subsist, and this term also includes his/her grandfather, grandmother in the case of a male and unmarried female employee, and her mother-in-law, father-in-law, in case of married female employee.

\end{solution}

\question What terms are defined in the Civil Service Act, 2049 ?

\begin{solution}

Following terms are defined in the act:

\begin{enumerate}
\item Civil service
\item Civil employee
\item Authority
\item Family
\item Head of department
\item Head of office
\item Permanent post
\item Expert post
\item Trade union of civil employees
\item Authentic trade union of civil employees
\item Prescribed
\end{enumerate}

\end{solution}

\end{questions}

### Classification of Civil Services

The provision relating to classification of civil services is prescribed under Article 3 of the act as "Constitution of Civil Service".

Government of Nepal shall constitute the civil services as follows:

- Nepal Economic Planning and Statistics Service
- Nepal Engineering Service
- Nepal Agricultural Service
- Nepal Judicial Service
- Nepal Foreign Service
- Nepal Administration Service
- Nepal Audit Service
- Nepal Forestry Service
- Nepal Miscellaneous Service
- Nepal Education Service

The service categories Nepal Health Service and Nepal Parliamentary Service were omitted by Nepal Health Act, 2053 (1997) and Parliamentary Secretariat Act, 2058 (2001), respectively.

### Classess in civil service (Article 4)

There are 4 classes in Gazetted and 5 classes in Non-gazetted ranks.

```{r classes-civil-service}
tribble(
  ~"Gazetted", ~"Non-gazetted",
  "Special", "First",
  "First", "Second",
  "Second", "Third",
  "Third", "Fourth",
  NA, "Fifth"
) %>% 
  knitr::kable(booktabs = TRUE, caption = "Classes in civil service") %>% 
  kableExtra::kable_styling(latex_options = "striped")
```

**Posts of service**

There shall be technical and administrative posts including the expert post in different classes of the civil service.

### Fulfillment of vacancy of civil service

Section 6A. Creation of position

- (Sub-Section 2, amended by Second Amendment) In preparing an organizational structure and creating a new position in order to establish a new governmental office or reviewing or altering the currently existing organizational structure and positions, the concerned Ministry shall carry out an organizational and management survey, based, inter alia, on the programme, work-load, job nature and available human resource, obtain consent of the Ministry of Finance and forward it to the Ministry of General Administration.
- (Subsection 2a) Based on the consent obtained pursuant to Sub-section 2, the Ministry of General Administration shall make its evaluation and submit it to the Government of Nepal, Council of Ministers for approval within Thirty days.

Provided that in the case of alteration in the organization structure and position, a report on such alteration in organization structure and position may be submitted after carrying out organization and management survey.

- (Subsection 3) Any post fulfilled by creating the position in contravention of Sub-section 1, 2 and 2a shall _ipso facto_ lapse.

Section 7. Fulfillment of vacancy in civil service

The posts of Gazette special class of the civil service shall be fulfilled pursuant to Section 19 and following posts shall be fulfilled as follows.

```{r vacancy-fulfilment}
tribble(
  ~"Post", ~"By open competition", ~"By evaluation of competency", ~"By internal competitive examination", ~"By adjustment of promotion by evaluation of work performance and experience", 
  "Classless/non-gazetted fifth class", 100, NA, NA, NA,
  "Non-gazetted fourth class", NA, NA, NA, NA,
  "Non-gazetted thrid class", NA, 100, NA, NA,
  "Non-gazetted second class", 60, 20, 20, NA,
  "Non-gazetted first class", 60, 20, NA, 20,
  "Gazetted third class", 70, 10, 10, 10,
  "Gazetted second class", 10, 60, 20, 10,
  "Gazetted first class", 10, 60, 20, 10
) %>% 
  knitr::kable(booktabs = TRUE, caption = "Fulfilment of Gazetted and Non-Gazetted posts of civil service") %>% 
  kableExtra::kable_styling(font_size = 8, latex_options = "striped") %>% 
  kableExtra::collapse_rows(1) %>% 
  kableExtra::column_spec(column = 1:5, width = c("12em", rep("8em", 4)))
```

- Out of posts to be filled up through internal competitive examination as referred to in Clause (d) of Sub-section (1), the non-gazetted second class posts shall be filled up through open competition from amongst the employees who are serving in non-gazetted third class and classless posts and have obtained minimum educational qualification and minimum service period.

- In order to make inclusive the civil service, 45% posts of the posts to be fulfilled by open competition shall be set aside and be filled up by having separate competition between the following candidates only, 

```{r quota-civil-service}
tribble(
  ~"Category", ~"Percentage", 
  "Women", 35,
  "Adiwasi/Janajati", 27,
  "Madhesi", 22,
  "Dalit", 9,
  "Disabled (differently abled)", 5,
  "Backwarded area", 4
) %>% 
  knitr::kable(booktabs = TRUE, caption = "Provisions for inclusive civil service (Sub-section 7, Section 7)") %>% 
  kableExtra::kable_styling(latex_options = "striped") %>% 
  kableExtra::add_footnote(label = "'Backwarded area' means Accham, Kalikot, Jajarkot, Jumla, Dolpa, Bajhang, Bajura, Mugu and Humla districts")
```

- (Sub-section 11, Section 7) The provision for the fulfillment of posts through the percent determined pursuant to Sub-section 7 shall be reviewed in every 10 years.
- (Sub-section 13, Section 7) Notwithstanding anything contained elsewhere in this Section (Section 7), the GoN may, by a notification in the National Gazette, specify the post that can be competed only by women for any specific nature of function or post.
- Notwithstanding anything contained elsewhere in this Section (Section 7), permanent appointment may be made by way of a competitive examination, between such women candidates only, specified by the Public Service Commission to the prescribed post for one time after the commencement of this Section as have obtained necessary minimum qualification for any post and have worked for at least five years temporarily in the permanent post of the women development program operated by a governmental body.

- (Sub-section 16, Section 7) The minimum qualification required for the post of gazetted second and first class to be filled up through open competition pursuant to Clauses (f) and (g) of Sub-section (1) shall be Master's Degree obtained on the concerned subject and experience of 5 years in the case of gazetted second class and that of 7 years in the case of gazetted first class, in an officer level post of any government service or a body corporate registered as per the prevailing law or an international association or institution.

Provided that, the term of work experience shall be less by 2 years in the case of a candidate who has done PhD.

- (Section 7A) Provisions on requisition for vacancy fulfillment. If any post of the civil service falls vacant in any manner, the concerned body shall give information thereof to the Public Service Commission within one month after such vacancy.

- (Section 7B) Restriction on vacancy fulfillment. 

(1) No post of the civil service shall be fulfilled in any manner other than that provided for in this Act.

(2) No salary shall be provided without having a salary report passed by the Civil Employee Records Office ( _Nijamati Kitabkhana_), in the case of civil employees serving in Ministries, Secretariats, Commissions and Departments and Offices situated within the Valley, and by the Treasury and Accounts Comptroller Office concerned, in the case of civil employees of the District based offices.

(3) (Includes provision for vindication on provisioning of salary in contravention to Sub-section 1, and the provision of salary to an employee holding a redundant post ( _Phajil_)).

- 'Employee holding a redundant post' means an employee whose organization or position has been repealed but his or her service is yet to be adjusted in another organization or service.

- (Section 9) Appointment to the post of civil service: Government of Nepal shall make appointment to the gazetted post of the civil service; and a Notification of such appointment shall be published in the Nepal Gazette.

The Head of Department may make appointment to the non-gazetted post of the civil service.

### Disqualifications for being candidate

The provisions relating to Disqualification for being candidate are described under Sub-section 1 of Section 10.

a. In the case of non-gazetted and classless post, those who have not completed the age of 18 years,
b. In the case of gazetted post, those who have not completed the age of 21 years,
c. In the case of male and female candidates, those who have already completed the age of 35 years and that of 40 years, respectively, provided that,

1. Any person who has not crossed the age of 40 years may be a candidate for such post of the civil service as may be specified for appointment of an ex-army or police thereto.

2. Any person who has not crossed the age of 45 years may be a candidate for the open competition of gazetted second and first class.

3. The age bar shall not be applicable in the case of permanent incumbent employees of the civil service.

d. Those who have been dismissed from service, with being disqualified for government service in the future,

e. Those who are non-Nepalese citizens,

f. Those who have been convicted by the court of a criminal offense involving moral turpitude.

- (Sub-section 2, Section 10) In the case of the female civil employee who has been temporarily appointed to any post of development side or development project of Government of Nepal and serving continuously since at least 5 years ago, the age bar shall not be applicable to her to become a candidate in the advertisement of the Public Service Commission for open competition.

- (Section 12) Retention of lien: The civil employee shall continue to have lien over his/her post in the following circumstances:

a. As long as he/she continues to function in that post;
b. Until the period of time allowed for the assumption of charge of another post, on being transferred to that post;
c. While he/she is on leave;
d. While he/she is under suspension;
e. While he/she is deputed by Government of Nepal on governmental or non-governmental business or while he/she is serving in any other post in an acting capacity.

- Section (13) Acting assignment

If any post of any Head of Department or Head of Office falls vacant or the employee holding such post goes on leave for a period exceeding 15 days or becomes subject to action pursuant to the prevailing law, a civil employee of the same grade or of one class lower post who has already become a potential candidate for promotion may be assigned to act in that post for a maximum period of 6 months, in order to perform the functions of that post for the time being. In cases where anyone has been so assigned to act in any post, that person or any other person shall not be again assigned to act in that post.

- (Section 17) Tenure of chief secretary and secretary shall be 3 and 5 years, respectively.

(...include other, refer to the section titles for overview. It should be taken that the body content is arcane.)

### Promotion (Section 19 and 20)

\begin{enumerate}
\item While making promotion to the gazetted post of the other civil services, except the gazetted post of Nepal Judicial Service, it shall be made on the recommendation of the promotion committee as referred to in Section 20.

\item Notwithstanding anything contained elsewhere in this Act, promotion to the post of gazetted special class shall be made as follows:

\begin{enumerate}
\item Government of Nepal shall, while making promotion to the post of Chief Secretary of the gazetted special class, make promotion by making selection on the basis of seniority and work efficiency from amongst the incumbent Secretaries.

\item The promotion committee shall, to the extent of availability, recommend those candidates who have secured the highest marks on the work performance evaluation, from amongst the potential candidates, in a number that is three times higher than the number of vacant posts, for promotion to the vacant posts of Secretary of gazetted special class or equivalent thereto; and Government of Nepal shall promote such candidate, from amongst the candidates so recommended, as it may hold appropriate on the basis of leadership capacity and work efficiency.

\item In order for any gazetted first class employee to be a potential candidate for the post of Secretary or special class post equivalent thereto, such employee must also have passed such a high level management training as prescribed by Government of Nepal.

\item In counting the work performance evaluation pursuant to Clause (b) for the promotion to the special class, the work performance evaluation of such years of minimum service period as required for promotion shall be counted.

\item Action for promotion shall be carried out twice in a year, One within the month of Poush and the other within the month of Ashad, each year.

\end{enumerate}

\end{enumerate}

**Promotion committee** (Section 20)

(Sub-section 1, Section 20) There shall be a promotion committee as follows to make recommendation for promotion to the other gazetted post of the civil service, except for promotion to the gazetted post of Nepal Judicial Service and for promotion from the non-gazetted post to the gazetted post:

1. Chairperson of the Public Service Commission or Member of the Public Service Commission designated by him/her -- Chairperson
2. Member of the Public Service Commission designated by the Chairperson of the Public Service Commission -- Member
3. Chief Secretary -- Member
4. Expert in the concerned subject nominated by the Public Service Commission -- Member
5. Secretary at the Ministry of General Administration -- Member secretary

(Sub-section 1a, Section 20) There shall be a promotion committee as follows, to make recommendation for promotion to the gazetted third class from the non-gazette first class:

\begin{enumerate}
\item Secretary at the Public Service Commission -- Member
\item Secretary at the Ministry of General Administration -- Member
\item Expert in the concerned subject nominated by the Chairperson of the Public Service Commission -- Member
\item Secretary at the Ministry relating to Service, Group or Sub-group or Gazetted First Class Officer designated by him/her -- Member secretary
\end{enumerate}

(Sub-section 1b, Section 20) The secretariat of the promotion committee referred to in Sub-section (1a) shall be at the Ministry operating the service, group or sub-group.

(Sub-section 2, Section 20) There shall be a promotion committee, as prescribed, to make recommendation for promotion to the non-gazetted post of the civil service.

(Sub-section 3, Section 20) The procedures to be followed by the promotion committee to make recommendation for promotion of the civil employees shall be as prescribed.

(Sub-section 4, Section 20) The promotion committee must, while publishing a promotion name list, also publish the marks obtained by the potential candidates who have filled up the forms for promotion.

(Sub-section 4a, Section 20) The promotion committee shall publish the results of promotion within four months from the last date of submission of application forms.

(Sub-section 5, Section 20) The promotion committee must prepare an annual plan of operations and carry out its functions as per that plan.

**Special provisions relating to promotion** (Section 20A)

\begin{enumerate}

\item Notwithstanding anything contained elsewhere in this Act, if any civil employee of the gazetted first class or below than that holding the same post for at least Fifteen years has One month left to get retirement pursuant to Section 33, such employee shall be promoted to one class higher post.

\item Notwithstanding anything contained elsewhere in this Act, if any civil employee of the gazetted first class or below than that holding the same post for at least Fifteen years intends to have voluntary retirement pursuant to Section 35 and makes an application therefor, such employee shall be granted retirement by promoting him/her to one class higher post.

\item A special post shall \textit{ipso facto} be created to make promotion pursuant to Sub-section (1) or (2) and that post shall \textit{ipso facto} lapse upon the retirement of such a civil employee from service.

\item Notwithstanding anything contained elsewhere in this Section, an employee who dies after the completing the period as referred to in this Section in the same post, the pension and other facilities payable to one level higher post shall be provided to the family of such an employee.

\item Notwithstanding anything contained elsewhere in this Section, any employees of gazetted first and second class holding the same post for at least fifteen years wish to take retirement with promotion, such employees shall be granted retirement by promoting them to one class higher post.

\end{enumerate}

**Minimum service period and educational qualifications required to be candidate for promotion** (Section 21)

\begin{enumerate}

\item In order to be a candidate for promotion, one must possess the educational qualifications as prescribed for the post that is One class below the class of the post to which promotion is made and have completed the minimum service period as follows:

\begin{enumerate}

\item Three years for a non-gazetted post,
\item Five years for a gazetted post

\end{enumerate}

\item Notwithstanding anything contained in Sub-section (1), even though any women, Adiwasi/Janjati, Madhesi, Dalit, Disable and permanently residing in the backward area civil employee has a service period that is the above-mentioned service period less one year, she shall be a potential candidate for promotion.

\end{enumerate}

**Ineligibility for begin potential candidate for promotion** (Section 22)

Notwithstanding anything contained in Section 21, a civil employee shall not be eligible to be a potential candidate for promotion in the following circumstances and during the following period:

\begin{enumerate}
\item In the event of suspension, during the period of suspension,

\item In the event of withholding of promotion, during the period of withholding,

\item In the event of withholding of increment in salary, during the period of withholding,

\item During the period of five years after the date of change in the service or group in the case of an employee who has so changed the service or group pursuant to Section 26 and after the change in the service or group in the case of an employee who has got such a change upon an application made by himself/herself.

\end{enumerate}

**Withholding and sanction of promotional appointment** (Section 22A)

If any civil employee is suspended or his/her appointment or salary increment is withheld after he/she has been recommended for promotion but before he/she gets promotional appointment, his/her promotional appointment shall then be withheld during the said period; and in the event of his/her acquittal of the suspension or after the expire of the period of withholding, he/she shall be granted appointment and only seniority shall be maintained as if he/she were not suspended or his/her promotion or salary increment were not withheld.

### Provisions for leave

```{r civil-service-leave-days}
tribble(
  ~"Leave type", ~"Frequency", ~"Remark", ~"Official empowered to approve",
  "Casual leave", "6 days", "Per year", "Concerned Chief of Office or Head of Department",
  "Festival leave", "6 days", "Per year", "Concerned Chief of Office or Head of Department",
  "Home leave", "1 day for every 12 work days", "An employee shall be entitled to accumulate upto 180 days of home leave earned by him/her", "Concerned Chief of Office or Head of Department",
  "Sick leave", "12 days", "Per year", "Concerned Chief of Office or Head of Department",
  "Maternity leave", "98 days prior to or after", "Each delivery", "Concerned Chief of Office or Head of Department",
  "Maternity care leave", "15 days prior to or after", "Each delivery", "Concerned Chief of Office or Head of Department",
  "Obsequies leave", "15 days", NA, "Concerned Chief of Office or Head of Department",
  "Study leave", "3 years maximum at once or various times", "With provision that GoN may add the upto 2 years period if considered necessary", "The GoN",
  "Extra-ordinary leave", "1 year at a time and upto 3 years during the service period", "To be entitled to the leave, an employee must complete 5 years of the service period (with exceptions). Leave is not to be counted in employees' service period.", "The GoN",
  "Leave without pay", "5 years maximum at one time or at several times", "Request made pursuant to Section 71A of the Civil Service act", NA
) %>% 
  knitr::kable(booktabs = TRUE, caption = "Provisions regarding leave in civil service", longtable = TRUE) %>% 
  kableExtra::kable_styling(latex_options = "striped", font_size = 8) %>% 
  kableExtra::column_spec(1:4, width = c("6em", "6em", "18em", "8em")) %>% 
  kableExtra::footnote(number = c(
    "Casual leave and festival leave may be taken for half a day. Both of these leave types cannot be accumulated and taken in the next year.", 
    "Maternity leave may be extended to maternity leave without pay for an additional period of 6 months in maximum as not to be deducted from any leave. A civil employee who has got the maternity leave and the maternity care leave shall be given an infant care allowance in a lump sum of Rs. 5000 for a maximum of 2 infants.",
    "To be entitled to get the study leave, an employee must have served in the service for 3 years.",
    "A civil employee going on the extra-ordinary leave shall not be entitled to salary during that period except when puruant to Sub-rule(10) of Rule 58, under which full salary will be provided.",
    "In cases where there falls a public holiday within the period of any type of leave taken by a civil employee, other than the casual and festival leave, home leave and sick leave, that period shall also be included in the leave taken by that employee.") %>% str_wrap(80), escape = FALSE)
```

