
# National Plans, Policies and Strategies

## Periodic planning

- The nineth plan considered poverty reduction and other development goals. The main target of this plan was to reduce the headcount poverty in the headcount poverty index from as high as 42% to 32%, 23%, 15% and 10 by 2002, 2007, 2012 and 2017, respectively. However, it could not be realized as the poverty rate in 2011 was 25%. 
- The tenth plan (2002-2007) is considered as a comprehensive plan designed for the reduction of poverty to a desired level. Designed as the poverty reduction strategy paper (PRSP), the tenth plan incorporated the involvement of the private sector in the market economy with the facilitation from the government side. Because of the bad political situation and existing conflict inside the country, reduction of poverty was a daunting task and this was reflected in the level of poverty that country was facing by the end of this plan.
- Instead of eleventh plan, interim plan for 3 years was launched from 2007/08-2009/10. The main objective of this plan was to generate an experience of a direct feeling of change in the lives of the general public by supporting in the establishment of peace and reducing the existing unemployment, poverty and inequality. This plan was designed in a critical state of transition where the country was formulating a new constitution to restore peace and reconstruction and reintegration among people suffering from the conflict.
- The twelve three year plans (2010/11-2012/13) has kept poverty alleviation as its major agenda and inclusive employment generation and equitable economic growth are considered two important pillars for realizing poverty alleviation. The purpose of this plan is to incorporate every community from all around the country to make sure that every underprivilaged and marginalized group is included. This plan has mobilized volunteers through National Development Volunteer Services (NDVS). NDVS has been implemented by district development committees. The purpose of this program is to mobilize volunteers as moderators in disadvantaged communities such as poor, ethnic, Dalit communities so that they will help them as facilitator in education, health, infrastructure, social mobilization and other areas.

## National planning commission

Introduction and History

The National Planning Commission (NPC) is the specialised and apex advisory body of the Government of Nepal for formulating a national vision, development policy, periodic plans and sectoral policies for overall development of the nation.
 
- It is headed by the Right Honorable Prime Minister. The NPC assesses resource needs, identifies sources of funding, and allocates budget for socio-economic development. 
- It serves as a central agency for monitoring and evaluating development policy, plans and programs. 
- The NPC also serves as an intellectual hub for the exchange of new development ideas and proposals from scholars, private sector, civil society, and development partners.

Government of Nepal (GoN) has also assigned following roles and responsibilities to NPC by NPC Executive Order  of March 12, 2018.

- To formulate basic development policies and prepare periodic development plans accordingly within the framework of a long-term development perspective.
- To explore internal and external resources along as well as indigenous and foreign technology and to recommend suggestions to GoN to accelerate the pace of development.
- To explore innovative approaches for sustainable development based on the economic situation of the country.
- To formulate annual programmes and assist GoN in their implementation.
- To advice GoN for the institutional development of monitoring & evaluation system, to monitor the implementation of programmes and projects according to plan target and outlay, and to carry out an imp`act evaluation of plans and projects.
- To provide guidelines, advice and suggestions to sectoral ministries, departments, other agencies and local bodies and assist them in the plans and projects formulation.
- To provide guidelines to collect data and to carry out action-oriented research necessary for the evaluation of new policies and for the refinement of the planning process.

Financial Administration Regulation (FAR) 1999 has assigned following roles and responsibilities to NPC:

- NPC has to provide guidelines on policy targets and priorities regarding formulation of budget for the coming fiscal year to the Ministry of Finance and all other line ministries before the issuance of circular on budget formulation.
- Formulate a policy framework for approval of the district level projects by the secretaries of the ministries.
- Approve Central level projects.
- Set policy targets on development programmes for the coming year within the month of December.
- Provide advice to GoN offices for accepting any kind of aid in kind or cash or reaching an agreement for foreign aid.

The National Planning Commission has been accomplishing various activities in order to carry out above mentioned roles and responsibilities. These activities are as follows:

- In addition to working as a national planning and programming body, the NPC has emerged as a leader and principle advisor in the economic development and economic affairs of the country. Almost all economic development and economic policies and programmes come into force either at the initiative of the Commission or on its advice or recommendations.
- NPC has emerged as a key coordinator in the management of economic development and general economic affairs of the country.
- NPC is also performing the role of an active and positive facilitator in the implementation of economic developmental policies and programmes. It has often assisted government organizations in solving challenges/problems faced in economic policy and programme implementation.
- The NPC has played the role of a pioneer in policy reforms. It has successfully pioneered the formulation and implementation of economic liberalization and other reform programmes.
- The Commission is working as the central level institution for monitoring and evaluation of economic development of the country.
- The Commission is a platform for exchange, dissemination, discussion and consultation in respect to economic affairs and development of the country. The meetings held in the offices of Vice-chairman and members of the Commission are important inputs for the development of the country.

The Commission also serves as an institution involved in analysis and finding solution to the problems of private sector in the country. The Policy Dialogue Committee under NPC is an effective forum for co-ordination of policy implementation and to address the various problems faced by the private sector

Membership

At present, the NPC has 
	- one full-time Vice-Chairman, 
	- six members, and 
	- one Member-Secretary who also heads a fully staffed secretariat.

The Chief-Secretary and the Finance Secretary are ex-officio members of the Commission. The national statistical organization of Nepal, the Central Bureau of Statistics (CBS), functions as a specialized entity of the NPC, headed by a Director-General.

The Commission consists of four functional divisions each headed by a Joint-Secretary: 

i. Economic Management; 
ii. Infrastructure and Production;
iii. Research and Evaluation; and 
iv. Good Governance and Social Development. 

Each division is further divided into Sections headed by an Under-Secretary.

Historical Background

The Planning Commission was first created in Nepal in 1956. It was soon renamed in accordance with the Yojana Mandal Act of 1957. Following the introduction of the partyless Panchayat system in 1961, the National Planning Council was formed under the then King. In 1963, the Council was dissolved and a new planning body, with an identical name, was constituted under the Chairman of the Council of Ministers. All the Ministers became ex-officio members of the Council; and the Ministry of Economic Affairs was renamed the Ministry of Economic Planning.

In 1968, all tasks related to development budget and foreign aid hitherto carried out by the Ministry of Economic Planning were assigned to the Ministry of Finance. The National Planning Council then morphed into the National Planning Commission (NPC) under the Chairmanship of the Rt. Hon'ble Prime Minister. A 1972 study on the functions and responsibilities of the central planning agency resulted in the reconstitution of the NPC.

After the historic restoration of multiparty democracy in 1990, the newly elected government reconstituted the NPC again with the Prime Minister as Chair, a full-time Vice-Chairman, five Members, and a Member-Secretary. The current organization and functions of the NPC draw on the Executive Order issued by the cabinet in 2018.

## National adaptation plan (NAP)

Nepal has adopted an approach to its NAP process that fits its specific context, taking into account the particular climatic and geographic characteristics of the country, opportunities and challenges associated with governance and the overarching development vision. This has yielded a number of unique features, notably:

- A process that will "leave no one behind": Inclusive NAP processing bringing in marginalized and disadvantaged communities, indigenous and traditional groups, with special consideration for youth, women and people with disabilities. Gender equality and social inclusion have been treated as both a cross-cutting issue and a standalone theme for the NAP process, providing a solid basis for addressing these issues in an informed and practical way, from teh beginning of the process.
  
- Integrated approach: In contrast with the sector- or ministry-based planning approach adopted by many countries, Nepal adopted an integrated approach, focusing on seven thematic areas, as well as two cross-cutting issues, with working groups focused on each. This has allowed for multidisciplinary dialogue and analysis throughout the process to date.

- Emphasis on stakeholder engagement: While the NAP process is led by the government, it has involved active participation of stakeholders from the very beginning, through the thematic and cross-cutting working groups. These groups bring together civil society organizations, academics and research organizations and private sector actors, among other representatives, enabling effective multistakeholder dialogue on the opportunities and gaps associated with adaptation within the different themes.

- An informed process: Recognizing the importance of technical and scientific rigour for adaptation planning, Nepal has made a considerable investment in research, analysis and technical capacity to inform the NAP process. This includes stocktaking reports and capacity gap assessments for the themes and cross-cutting issues, analysis of climate trends and future scenarios and the development of a comprehensive framework for vulnerability and risk assessment. Collectively, these efforts provide a solid base of information and analysis upon which informed decisions can be made regarding adaptation planning and implementation.

An important issues to the context in NAP implementation is the clarification of the role of sub-national actors and the creation of strategic linkages between national and sub-national adaptation planning as the new governance systems and structures are established. Related to this is the challenge of ensuring that the process is gender-balanced and inclusive, in line with the commitment to leave no one behind.

Nepal's NAP process builds on past experience with adaptation planning, including through the National Adaptation Programme of Action (NAPA), developed in 2010, and the Framework on Local Adaptation Plans for Action (LAPA), developed in 2011, which has facilitated development of adaptation plans by Village Development Committees across the country. The government, led by the Ministry of Population and Environment has integrated the lessons learned from these past efforts in the NAP process, taking a deliberate and thoughtful approach that aims to be participatory and inclusive, while integrating the best available scientific information about current and future climate risks.

## National Climate Change Policy (NCCP), 2076

Reasons for Nepal's vulnerability to climate change are Topographical fragility, fragile geological structure, sensitive ecosystems and diversity of climate and micro-climate zones prevalence in Nepal. Additionally, poverty, illiteracy, social disparity and high dependence of communities on natural resources for livelihood.

- Nepal National Red Strategy implemented to check greenhouse gases emmission from deforestation and forest destruction.
- Provision of climate change coordination committee in Chairmanship of Prime minister.
- Establishment of Climate change management section under Ministry of Forest and Environment as focal point for implementation of international treaty.
- Provision of Climate budget code by National planning commission.
- Alternative energy promotion center has established climate and carbon unit for executing Clean Development Mechanism (CDM) related activities.

**Goal**

To contribute to socio-economic prosperity of the nation by building a climate resilient society.

**Objectives**

a. To enhance climate change adaptation capacity of persons, families, groups and communities vulnerable to, and at risk of, climate change;
b. To build resilience of ecosystems that are at risk of adverse impacts of climate change;
c. To promote green economy by adopting the concept of low carbon emmission development;
d. To mobilize national and international financial resources for climate change mitigation and adaptation in just manner;
e. To conduct research, make effective technology development and information service delivery related to climate change;
f. To mainstream or integrate climate change issues into policies, strategies, plans and programs at all levels of State and sectoral areas;
g. To mainstream gender equality and social inclusion (GESI) into climate change mitigation and adaptation programs.

**Policies** (Sectoral policy, strategies and working policies)

1. Agriculture and food security
2. Forest, biodiversity and watershed conservation
3. Water resources and energy
4. Rural and urban habitats
5. Industry, transport and physical infrastructure
6. Tourism and natural and cultural heritage
7. Health and drinking water and sanitation
8. Disaster risk reduction and management
9. Gender equality and social inclusion, livelihoods and good governance
10. Awareness raising and capacity development
11. Research, technology development and expansion
12. Climate finance management

Policy number 9-12 are inter-thematic areas.

## Medium term expenditure framework (MTEF)

Recognizing the fiscal deterioration and the need to revise the Tenth Plan accordingly, the Government initiated the preparation in parallel of a Medium Term Expenditure Framework (MTEF) in late 2001. The key objective of the MTEF was to begin to implement the Tenth Plan from the beginning of the fiscal year 2002/03, without waiting for the formal finalization of the Plan. Accordingly, the key projects/programs and activities that are considered essential for achieving the Tenth Plan's poverty reduction goals and their resource requirements for the next three years were to be identified, and were to be given priority in terms of budget allocations in the 2002/03 budget. Additional considerations guiding the MTEF, in view of the fragility of the fiscal situation, were: (i) How to streamline the budget which was already overextended well beyond resource availability, with over 600 ongoing activities/budget lines ? (ii) How to protect the Tenth Plan's key priorities in the event of further shortfalls in resource availability ? and (iii) How to ensure that the resources allocated and released for priority activities would be effectively utilized? The MTEF was expected to be the principal instrument for operationalizing the Tenth Plan, by prioritizing the proposed Plan activities according to the changing resource situation, and firmly linking the annual budget process and the Plan.

Thus, a three year MTEF was prepared in early 2002 by virtually the same line ministry teams responsible for Tenth Plan preparation, using similar methodology for ranking and prioritizing activities, (but with a sharper focus on implementability, short term financing needs and results). On the basis of the MTEF, (i) the 2002/03 development budget was initially set at a significantly lower level -- Rs. 38 billion, compared to Rs. 50 billion in the 2001/02 budget; and the number of budget lines were reduced from over 600 to 430. (ii) Even more importantly, all activities were classified into three groups according to their priority (P1being the highest priority and P3 being the lowest); and (iii) The principle was established that the priority classification would be strictly followed in releasing funds for development activities. Thus, P1s will receive first priority in budget releases; and P2 and P3 will get funded only if funds are still available after providing for P1s. This has established clear criteria and methodology for adjusting resource allocations on the basis of Tenth Plan's priorities, in line with changes in the resource situation. (iv) Finally, resource allocation was linked to expenditure reporting and performance, in order to ensure satisfactory results. Thus, under the new system that has been evolved, activities which do not provide statements of expenditures (SOEs) for the preceding trimester will not be given additional funds, until they provide the SOE's.

## 10th plan (FY: 2059/60-2063/64)

Four pillars of poverty reduction strategy:

1. Broad based economic growth; 
2. Social sector development including human development;
3. Targeted programs including social inclusion, in order to bring the poor and marginalized groups into the mainstream of development, together with targeted programs for the ultra poor, vulnerable and deprived groups (who may not adequately benefit from the first two pillars); and 
4. Good governance.

The plan stresses strategic cross-cutting approaches with regard to: 

(a) redefining the role of the State, and limiting public interventions; 
(b) enlisting the private sector to play a leading role in employment and income generation and together with NGOs, INGOs and CBOs, in complementing government efforts in service delivery functions in key areas, as well as in implementing key activities; 
(c) promoting community participation in and management of activities at the local levels; and 
(d) accelerating the decentralization process, which is also a key element under good governance

Implementation strategy of four-pillars and their implications for sectoral programs and activities are:

1. Macroeconomic framework
2. Structural reform agenda
3. Broad based economic growth
4. Social sector development
5. Social inclusion and targeted programs
6. Good governance
7. Improving financial management and accountability
8. Decentralization

```{r tenth-periodic-plan}
read_csv("./data/periodic_plan_10th.csv", skip = 1) %>% 
  rename_all(function(x)str_replace_all(x, "_", " ") %>% str_to_sentence()) %>% 
  knitr::kable(booktabs = TRUE, caption = "Achievements of 9th and goals for 10th periodic plan under normal and restrained cases.", longtable = TRUE) %>% 
  kableExtra::kable_styling(latex_options = "striped", font_size = 8) %>% 
  kableExtra::collapse_rows(columns = 1, valign = "middle") %>% 
  kableExtra::column_spec(column = 1:5, width = c("10em", "10em", "8em", "8em", "8em"))
```


**Decentralization as implementation strategy of 10th plan**

Decentralization is an important means of bringing development closer to the rural poor—by involving local communities in developing appropriate programs which are best suited to their needs and in implementing them, ensuring greater accountability for use of public resources, and mainstreaming the poor and deprived groups. The local Self Governance Act (LSGA, 1999) provides a framework for decentralization, which the government has agreed to implement in a phased manner. The Local Bodies Finance Commission (LBFC) has also recently made recommendations for fiscal devolution. A Decentralization Implementation and Monitoring Committee was also set up to oversee effective implementation. But progress so far has been hampered by institutional capacity and fiscal constraints, and more recently by the dissolution of elected local bodies.

The main objectives of the Tenth Plan in this area are to ensure greater participation of people in the governance process to accelerate the development process by implementing fiscal devolution in a phase-wise manner as envisaged in the LSGA; and creating necessary institutional mechanisms including the formation of a Local Service Commission. The main strategies being adopted to achieve these objectives include: the devolution of basic service delivery functions, capacity building of local bodies and decentralization of certain revenue mobilization functions to local bodies.The major activities being undertaken to carry out this strategy include the transfer of the management and operation of DIDO, schools, health posts, postal service, agriculture and small irrigation projects, as well as rural roads, to local bodies. The government will constitute local service cadres and enhance the planning, management and evaluation capacities of the local bodies by providing the necessary training and equipment. Revenue collection and fiscal management of local bodies will be improved and auditing and accounting system refined. As a result of these initiatives, it is expected that over the Plan period, 25,000 primary schools, all sub-health posts and agriculture extension services will be transferred to local bodies. The share of local revenue in the local bodies' budget will rise, transparency and accountability will increase and service delivery to communities at village levels will improve considerably.

## 14th plan (FY: 2073/74-2075/76)

- Recently proposed projects: 13
- 21.6% of the population is below poverty line by the end of 13th plan

```{r forteenth-periodic-plan}
read_csv("./data/periodic_plan_13th.csv", skip = 1) %>% 
  knitr::kable(booktabs = TRUE, caption = "Goals and achievements of 13th periodic plan (2073/74-75/76) for selected indicators") %>% 
  kableExtra::kable_styling(latex_options = "striped") %>% 
  kableExtra::column_spec(column = 1, width = "18em")
```

- Gross national income in 13th periodic plan was expected to rise by 6.0 percentage, however by 2070/71 it was 5.72%, in 2071.72 2.32% and in 2072/73 it remained 0.77%. Hence resulting in average economic growth of 2.92%.
- There was a projection for an increase in 4.5% in agriculture sector, however, in years 2070/71, 71/72 and 72/73 respectively, 4.55%, 0.81% and 1.33% growth were achieved. Hence resulting in average growth of 2.22%
  - Food grain production was projected to reach 1,08,81,000 mt annually. However at the second year of the plan only 92,66,000 mt were realized.
  - There was an estimation of 12,60,000 mt production of fruits. But only 11,86,000 mt were produced.
  - There was an estimation of 36,00,000 mt production of vegetables. However, 36,23,000 mt were produced of it, surpassing the set target.

## 15th plan
    
\begin{questions}

\question Describe the 15th plan approach paper:
% \makeemptybox{1.5in} % use this to provide a box in question

\begin{solution}

Objectives

\begin{itemize}
\item Development of affordable, quality and modern infrastructure. Increment of productive, dignified employment opportunities. High, sustainable and comprehensive economic growth and poverty alleviation to lay foundations of prosperity.
\item Quality health and education, healthy and balanced environmental conditions, social justice and accountable public service will improve federal governance. Citizens will be realize improved and dignified ways of life.
\item Socio-economic transformation and development of sovereign national economy thereby resulting in a protection of national welfare, sovereignty and independence.
\end{itemize}

Strategies

\begin{itemize}
\item High, sustainable and employment generating economic growth
\item Affordable and qualitative health and education service assurance
\item Domestic and international interconnectedness and sustainable urban infrastructure development
\item Increase production and productivity
\item Provisioning of complete, sustainable and productive social security and protection. 
\item Poverty alleviation and economic and creation of judicious social system that ensures social equality.
\item Conservation of natural resources, it's mobilization and promotion.
\item Strengthening of public service, maintaining balance at provincial level and foster national unity.

\end{itemize}

\end{solution}

\end{questions}
