# Technology, communication and health

## Technology and energy

\begin{questions}

\question (Long answer type) What is energy crisis ? What are the measures that can be taken to solve the crisis ? What are the different sectors affected by energy crisis ?

\question Explain briefly on the prospect of hydroelectricity development in Nepal.

\end{questions}

## Health

### Disease

A disease is a particular abnormal condition that negatively affects the structure or function of all or part of an organism, and that is not due to any immediate external injury.

Diseases are often known to be medical conditions that are associated with specific symptoms and signs. A disease may be caused by external factors such as pathogens or by internal dysfunctions. For example, internal dysfunctions of the immune system can produce a variety of different diseases, including various forms of immunodeficiency, hypersensitivity, allergies and autoimmune disorders.

In humans, disease is often used more broadly to refer to any condition that causes pain, dysfunction, distress, social problems, or death to the person afflicted, or similar problems for those in contact with the person.

There are four main types of disease: infectious diseases, deficiency diseases, hereditary diseases (including both genetic diseases and non-genetic hereditary diseases), and physiological diseases. 

Diseases can also be classified in other ways, such as communicable versus non-communicable diseases.

1. Communicable or infectious diseases: These are caused by a variety of pathogenic viruses, bacteria, protozoa, fungi and worms. The pathogens are generally carried with the help of a vector.

2. Non-communicable or non-infectious or degenerative diseases: These occur due to the malfunctioning of some organ or organ system in the body. The type of disease is defined as:

"Chronic condition that do not result from an (acute) infectious process and hence are "not communicable".


```{r human-diseases}
readxl::read_xlsx("./data/health_human_diseases_infectious.xlsx", sheet = "bacteria-virus", skip = 1) %>% 
  knitr::kable(caption = "Bacterial and viral diseases of humans", booktabs = TRUE, longtable = TRUE) %>% 
  kableExtra::kable_styling(latex_options = "striped") %>% 
  kableExtra::collapse_rows(columns = 1) %>% 
  kableExtra::column_spec(column = 1:6, width = c("5em", "6em", "6em", "8em", "8em", "6em"))

readxl::read_xlsx("./data/health_human_diseases_infectious.xlsx", sheet = "fungi-nematode") %>% 
  knitr::kable(caption = "Fungal and nematode diseases of humans", booktabs = TRUE, longtable = TRUE) %>% 
  kableExtra::kable_styling(latex_options = "striped") %>% 
  kableExtra::collapse_rows(columns = 1) %>% 
  kableExtra::column_spec(column = 1:6, width = c("5em", "6em", "6em", "8em", "8em", "6em"))
```


**Non communicable diseases**

- Non-communicable diseases (NCDs) include cardiovascular disease, diabetes, cancer, chronic lung diseases, genetic diseases, allery and nutritional deficiencies -- micronutrients, vitamins.
- People whose health is compromised by NCDs are more vulnerable to the stresses and disruptions of disaster
- Risk reduction of NCDs requires multisectoral action to promote healthy behaviors. NCDs are managed by a range of health services that address prevention, treatment, rehabilitation and palliative care
- Health infrastructure with resilience to disasters is necessary to ensure continuity of care for people with NCDs in emergencies. Specific measures may include protection of essential equipment and stockpiling of essential medicines.
- A significant number of deaths after a disaster are as a result of inadequate health care services to cater for pre-existing conditions and illnessess.

Characteristics of NCDs

- Complex etiology (causes)
- Multiple risk factors
- Long latency period
- Non-contagion origin (non communicable)
- Prolonged course of illenss
- Functional impairment of disability

Types of NCDs

- Cardiovascular disease (e.g., Coronary heart disease, stroke)
- Cancer: 7.6 million people died from cancer in 2008 and the deaths from this disease is estimated to reach 13.1 million by 2030. Breast cancer, prostate cancer and colorectal cancer are successively the most common types of cancer.
- Chronic respiratory disease
- Diabetes
- Chronic neurologic disorders (e.g. Alzheimer's dementias)
- Arthritis/Musculoskeletal disease
- Nutritional disorder
- Unintentional injuries (e.g. from traffic crashes)

Modifiable risk factors of NCDs

- Physical inactivity: Increases risk of type-2 diabetes, certain cancers, heart attacks, stroke, falls and early death.
- Tobacco use: Tobacco kills upto half of its users; globally 6 million people die each year. Following disease are common among smokers: Cancer, coronary heart disease, disease of lungs, peripheral vascular disease, stroke, fetal complications and stillbirth
- Alcohol use: 11.5% of all global drinkers are episodic, heavy users. Globally, 2.5 million people die from alcohol consumption per year. Has long term effects of liver diseases, cancers, hypertension, gastrointestinal disorders, neurological issues and psychiatric issues.
- Unhealthy diets (increased fat and sodium, with low fruit and vegetable intake)

Non-modifiable risk factors

- Age
- Gender
- Race
- Family history

Metabolic risk factors are generally modifiable risk factors refering to biochemical processes involved in the body's normal functioning. They include:

- Raised blood pressure

```{r blood-pressure-measure}
tribble(~"Measurement", ~"Normal", ~"Pre-hypertensive", ~"Hypertensive", 
        "Systolic (mgHg)", "<120", "120-139", "140+", 
        "Diastolic (mmHg)", "<80", "80-90", "90+") %>% 
  knitr::kable(caption = "Standardized blood pressure measurements", booktabs = TRUE) %>% 
  kableExtra::kable_styling(latex_options = "striped")

```

- Raise total cholesterol: Increases the risk of heart disease and stroke. A 10% reduction in serum cholesterol in men aged 40 has been reported to result in a 50% reduction in heart disease within 5 years.
- Elevated glucose
- Overweight and obesity: Abnormal or excessive fat accumulation that presents a risk to health. A Body Mass Index (BMI) of 30 or more indicated obesity in $BMI = \frac{\text{Weight in kg}}{\text{Height in meters}^2}$


Disease spread through blood transfusion

AIDS (Acquired Immuno Deficiency Syndrome): It destroys immune system of the body & is caused by the Human Immuno deficiency Virus (HIV).There are two types of HIV namely HIV-1 and HIV-2. The most common virus currently associated with AIDS is HIV-1. A virus found in the blood of wild African green monkey called the Simian Immuno deficiency Virus (SIV) is similar to HIV-2. HIV is a retrovirus. It can synthesize DNA from RNA. The major cell infected by HIV is the helper T-Iymphocyte that bears the CD-4 receptor site. HIV progressively destroys T-Iymphocytes. The patient occasionally will suffer from swollen lymph nodes, mild prolonged fever, diarrhoea or other non-specific symptoms.

HIV antibodies can be detected by the ELISA test (Enzyme-Linked Immuno Sorbent Assay). World AIDS Day is celebrated on 1st December.

### Nutritional deficiencies

Different types of vitamins and minerals enable healthy body function, such as cell and tissue repair, production of cells, and healthy brain function. Your body doesn't produce these vitamins and minerals on its own, and while you can get most of these from a supplement, your body reacts better when these vitamins and minerals come from food.

Water-soluble vitamins

Thiamine (Vitamine B1) deficiency is especially common in countries that do not require fortification of wheat and maize flour and rice to replace the naturally occurring thiamine content lost to milling, bleaching and other processing. Severe deficiency causes beriberi, which became prevalent in Asia as more people adopted a diet primarily of white rice. Wernicke encephalopathy and Korsakoff syndrome are forms of beriberi. Alcoholism can also cause vitamin deficiency. Symptoms of deficiency include weight loss, emotional disturbances, impaired sensory perception, weakness and pain in the limbs, and periods of irregular heart beat. Long-term deficiencies can be life-threatening. Deficiency is assessed by red blood cell status and urinary output.

Riboflavin (Vitamine B2) deficiency is especially common in countries that do not require fortification of wheat and maize flour and rice to replace the naturally occurring riboflavin lost during processing. Deficiency causes painful red tongue with sore throat, chapped and cracked lips, and inflammation at the corners of the mouth (angular cheilitis). Eyes can be itchy, watery, bloodshot and sensitive to light. Riboflavin deficiency also causes anemia with red blood cells that are normal in size and hemoglobin content, but reduced in number. This is distinct from anemia caused by deficiency of folic acid or vitamin B12, which cause anemia.

Niacin (Vitamine B3) deficiency causes pellagra, a reversible nutritional wasting disease characterized by four classic symptoms often referred to as the four Ds: diarrhea, dermatitis, dementia, and death. The dermatitis occurs on areas of skin exposed to sunlight, such as backs of hands and neck. Niacin deficiency is a consequence of a diet low in both niacin and the amino acid tryptophan, a precursor for the vitamin. Chronic alcoholism is a contributing risk factor. Low plasma tryptophan is a non-specific indicator, meaning it can have other causes. The signs and symptoms of niacin deficiency start to revert within days of oral supplementation with large amounts of the vitamin.
Pantothenic acid (Vitamine B5) deficiency is extremely rare. Symptoms include irritability, fatigue, and apathy.

Vitamin B6 deficiency is uncommon, although it may be observed in certain conditions, such as end-stage kidney diseases or malabsorption syndromes, such as celiac disease, Crohn disease or ulcerative colitis. Signs and symptoms include microcytic anemia, electroencephalographic abnormalities, dermatitis, depression and confusion.

Biotin (Vitamin B7) deficiency is rare, although biotin status can be compromised in alcoholics and during pregnancy and breastfeeding. Decreased urinary excretion of biotin and increased urinary excretion of 3-hydroxyisovaleric acid are better indicators of biotin deficiency than concentration in the blood. Deficiency affects hair growth and skin health.

Folate (Vitamin B9) deficiency is common, and associated with numerous health problems, but primarily with neural tube defects (NTDs) in infants when the mother's plasma concentrations were low during the first third of pregnancies. Government-mandated fortification of foods with folic acid has reduced the incidence of NTDs by 25% to 50% in more than 60 countries using such fortification. Deficiency can also result from rare genetic factors, such as mutations in the MTHFR gene that lead to compromised folate metabolism. Cerebral folate deficiency is a rare condition in which concentrations of folate are low in the brain despite being normal in the blood.

Vitamin B12 deficiency can lead to pernicious anemia, megaloblastic anemia, subacute combined degeneration of spinal cord, and methylmalonic acidemia, among other conditions. Supplementation with folate can mask vitamin B12 deficiency.

Vitamin C deficiency is rare. Consequently, no countries fortify foods as a means of preventing this deficiency. The historic importance of vitamin C deficiency relates to occurrence on long sea-going voyages, when the ship food supplies had no good source of the vitamin. Deficiency results in scurvy when plasma concentrations fall below 0.2 mg/dL, whereas the normal plasma concentration range is 0.4 to 1.5 mg/dL. Deficiency leads to weakness, weight loss and general aches and pains. Longer-term depletion affects connective tissues, severe gum disease, and bleeding from the skin.

Fat-soluble vitamins

Vitamin A deficiency can cause nyctalopia (night blindness) and keratomalacia, the latter leading to permanent blindness if not treated. It is the leading cause of preventable childhood blindness, afflicting 250,000 to 500,000 malnourished children in the developing world each year, about half of whom die within a year of becoming blind, as vitamin A deficiency also weakens the immune system. The normal range is 30 to 65 microg/dL, but plasma concentrations within the range are not a good indicator of a pending deficiency because the normal range is sustained until liver storage is depleted. After that happens, plasma retinol concentration falls to lower than 20 microg/dL, signifying a state of vitamin A inadequacy.

Vitamin D deficiency is common. Most foods do not contain vitamin D, indicating that a deficiency will occur unless people get sunlight exposure or eat manufactured foods purposely fortified with vitamin D. It is typically diagnosed by measuring the concentration of the 25-hydroxyvitamin D (25(OH)D) in plasma, which is the most accurate measure of stores of vitamin D in the body. Deficiency is defined as less than 10 ng/mL, and insufficiency in the range of 10-30 ng/mL. Serum 25(OH)D concentrations above 30 ng/mL are "not consistently associated with increased benefit." Serum concentrations above 50 ng/mL may be cause for concern. Vitamin D deficiency is a known cause of rickets, and has been linked to numerous other health problems.

Vitamin E deficiency is rare, occurring as a consequence of abnormalities in dietary fat absorption or metabolism, such as a defect in the alpha-tocopherol transport protein, rather than from a diet low in vitamin E. The US Institute of Medicine defines deficiency as a blood concentration of less than 12 micromol/L. Deficiency causes poor conduction of electrical impulses along nerves due to changes in nerve membrane structure and function.

Vitamin K deficiency as a consequence of low dietary intake is rare. A deficient state can be a result of fat malabsorption diseases. Signs and symptoms can include sensitivity to bruising, bleeding gums, nosebleeds, and heavy menstrual bleeding in women. Newborn infants are a special case. Plasma vitamin K is low at birth, even if the mother is supplemented during pregnancy, because the vitamin is not transported across the placenta. Vitamin K deficiency bleeding (VKDB) due to physiologically low vitamin K plasma concentrations is a serious risk for premature and term newborn and young infants. Untreated, consequences can cause brain damage or death. The prevalence of VKDB is reported at 0.25 to 1.7%, with higher risk in Asian populations. The recommended prevention treatment is an intramuscular injection of 1 mg of vitamin K at birth (called the Vitamin K shot.) There are protocols for oral administration, but intramuscular injection is preferred.

\begin{questions}

\question What are the causes of malnutrition ?

\begin{solution}

Malnutrition is a condition that results from eating a diet in which one or more nutrients are either not enough or are too much such that the diet causes health problems. Every year 406,000 people die globally due to malnutrition. In economically impoverished country like Nepal, majority cases of health related issues arise due to undernutrition, rather than due to excess of nutrition. Although, it is touted that Nepal is an agricultural country and its plains a storehouse of granary, prevalance of nutritional deficiency (an important factor setting back Nepal's food security achievement) is staggering. Following are the main factors contributing to malnutrition:

\begin{enumerate}
\item Lack of nutritious diet
\item General public lack awareness on importance of nutritious diet
\item Lack of skills and techniques on cooking and food storage with maximum nutrient retention
\item Inability of properly constituting a diet despite availavility of range of food items
\item Lack of proper installments for safe storate of food
\item Due to lack of suitable supply system for remote hilly and mountainous regions, general scarcity of adequate food is observed
\item Lacking awareness on general importance of food
\item Campaigns aimed at awaring the public have not been effective.
\end{enumerate}
\end{solution}

\question Discuss about the importance of environment on human health.

\begin{solution}
About one quarter of the diseases worldwide can be accredited to environmental causes, with 40\% of the sufferers being children under five.

Among the ten most significant dangers to the developing countries with high-mortality rate are unsafe water and sanitation and hygiene problems. Just about 80 per cent of each and every type of infectious disease is caused by unclean water. Globally, 1.1 billion people lack access to safe water supply and 2.4 billion to adequate sanitation.

In East Asia and the Pacific, 48 per cent of the population have access to improved hygiene while in South Asia, merely 37 per cent have such access. Inadequate water and insufficient hygiene lead to ill health.

Over 200 million people in nine Asian countries, more often than not living in rural areas, are at risk of arsenicosis, cancer and eventual death from the long-term use of groundwater infected with arsenic.

Arsenic toxicity has a pervasive effect on all systems in the body and long-term consequences on children, slowing their cognitive expansion and cerebral ability. There is no known medical cure for arsenicosis. The only sure way of preventing arsenicosis is to avoid drinking impure water. Furthermore, extreme fluoride in water causes dental and skeletal fluorosis. Lead emissions are also a serious problem in the region.

In the Asian and Pacific area, indoor air contamination poses the uppermost health risk. Contact with smoke from solid fuel like wood, dung, coal and agricultural remains increases the amount of contamination. Indoor air pollution causes an accumulation of pollutants, such as coarse and fine particulate matter, carbon monoxide, oxides of nitrogen, benzene, toluene and formaldehyde, in the bodies of those exposed to indoor air pollution. Epidemiological evidence suggests a link between indoor air toxic waste and respiratory infections, TB, cancers, low birth weight and greater infant and prenatal mortality. 

Nepal's performance seems to be worst when it comes to protecting environment and combating air pollution. Nepal is among the bottom five countries on the Environmental Performance Index 2018, plummeting 27 points from 149 in 2016, according to a biennial report by Yale and Columbia Universities along with the World Economic Forum. Air pollution in Nepal, particularly in the Kathmandu Valley, has emerged as a serious environmental health risk.

Rapid urbanisation, unplanned development activities, smoke from brick kilns, ongoing road widening drive and construction works for the Melamchi Water Supply project and increasing number of vehicles have contributed in making Kathmandu one of the most polluted cities in the world.

Furthermore, the concentration of particulate matter is observed higher during night in winter because of the formation of inversion layer which acts as the blanket to trap pollutants. The Particulate Matter (PM2.5) concentration in Kathmandu is five times more than the National Ambient Air Quality Standard of Nepal set by the government, and way more than the World Health Organisation (WHO) standard.

Most of air pollution-related deaths are from non-communicable diseases and in terms of global disease burden, air pollution is the cause of over one-third deaths from stroke, lung cancer and chronic respiratory diseases, says the WHO. One-quarter of deaths occur due to ischemic heart disease. According to a WHO report, nearly 740 people died from acute lower respiratory infection, 1,770 from chronic obstructive pulmonary disorder, 932 from lung cancer, 3,328 from ischemic heart disease and 3,183 from stroke in Nepal in 2016.

However, the number of vehicles being registered in Kathmandu is rapidly increasing. Weak institutions, corruption and lack of technology are also adding to Kathmandu’s air pollution. One of the key issues is monitoring of air quality by using standard, good quality and practically identical strategies and instruments.

WHO points to a lack of reliable public transport system. Transport system needs to be strengthened and alternative mode of transport such as cycling for short routes must be encouraged by creating safe cycling routes. In Kathmandu, 41 per cent of people travel by foot while 2 percent people use bicycles. So, walking and cycling should be made more convenient.

Apart from its instant toxic and carcinogenic effects, air contamination affect the environment and health in other ways. Some air pollutants descend as acid rain. Acid rain causes ecological degradation which impacts ecosystems, increasing health hazards. 

Ozone depletion, associated with use of ozone-depleting chemicals, is expected to augment UV-B radiation, resulting in the oppression of immune systems and skin cancer. In South East Asia, 70 million people are seasonally affected by air pollution due to forest fire.

A number of pesticides are associated with cancers, miscarriages and inborn malformations. Agrochemical poisoning and agro-machinery accidents cause diseases, injuries and death that are underreported.

Small-scale farmers are especially vulnerable to the exposure to chemical fertilisers and pesticides that are widely used in the region, and to equipment-related hazards aggravated by sustained hard physical labour with old and poorly maintained gear.

A lot of new infectious diseases have emerged because viruses have jumped the species fence from flora and fauna to humans. Undeveloped production systems, in particular intensive animal husbandry in densely populated areas, with livestock in close proximity to humans, are the perfect breeding ground for new diseases.

SARS as well as avian influenza are fresh examples. These diseases affect not only humans, but also flora and fauna and vegetation. They also threaten food safety and livelihood. Thus, these diseases have themselves turned out to be drivers of environmental alteration.

A notorious issue in foodstuff production is the use of genetically modified organisms. There is vigorous global debate on the large-scale use of genetically modified organisms in developed countries when their long-term penalty on human health has not been well-known.

Plastic is extensively used for wrapping food. These substances could be released if ingested. Vast quantities of microscopic plastic fragments are building up in oceans. The accumulation of plastic wreckage in parts of the central Pacific Ocean is six times greater than that of resident plankton, the very foundation of the food chain. There is indication that plastic debris is being ingested by marine organisms and might go into the food chain.

Haphazard waste disposal is a chief health danger. Non-biodegradable plastic waste constitutes a main part of waste. Plastic floats in water; it accumulates and absorbs toxic hydrophobic chemicals from other sources. Artificial waste also clogs sewerage systems, causing severe hygiene problems.

In spite of some progresses made since the Basel Convention on the Trans Boundary Movement of Hazardous Waste came into force, large amounts of hazardous waste, including electronic waste, are still exported to developing countries in the region. Recycling is often absent and the burning of plastic and dumping are common.

The unsafe disposal of hazardous domestic medical and manufacturing waste pose severe health risks to people in the region. Governments in the region should pay serious attention to ecological health.

\end{solution}

\end{questions}

## Media and information communication technology

### Role of Media

There is no denying the fact that media play a significant role in national affairs and national development. It not only promotes national identity and harmony but also contributes in achieving economic progress and social development. As the media touches almost every aspect of public life, its responsibility is crucial for creating, shaping and reflecting public opinion. Considered as the heart of democracy and rule of law, the media also helps nations meet various formidable contemporary challenges. Nepal has witnessed a noteworthy growth in the media sector following the restoration of democracy in 1990. The Rising Nepal's sister publication, the Gorkhapatra daily, has been a national heritage in the country's media sector. The vernacular Nepali newspaper has entered the 120th year of its launching. As a state media, the publication has seen many ups and downs since its inception on May 6, 1901 (Baisakh 24, 1958 B.S.). Despite facing a myriad of challenges, the publication has continued to serve the nation and the people.

Human life is regulated by communication of which media constitute a part, a truth producing part. It is a process of learning, awareness and coordination of behaviour. It helps to visualise the life's prospects. Media make up a union of soft power of text, image and sound to get the voice of people into the public sphere and widen the scale of their deliberation and participation in decisions. In the multi-sided process of nation building various insights, perspectives and concepts of media are useful resources to satisfy peoples' yearning for freedom, justice and national identity. Media thus offer a marketplace of ideas to exercise choice. Their autonomy in rules, finance, professionalism and context-sensitivity enable them to freely report about the vices of Nepali society and offer options to mitigate them.

Nation building under the state's leadership and mobilisation of civic bodies with robust technical and financial support are the most desirable gifts in a world of disarray. Modern means of communication, cultural diffusion, justice and democracy are propelling the fusion of Nepali nations into its state symbolising an epithet of public will. Now, Nepali leaders need to transmit specific interests of groups into public and national interest and socialise them to identify with Nepaliness. It is not by the secluded tribal identity but equal citizenship that defines the Nepali state and its civic nationalism defies predators.

Its impersonal spirit can abolish feudalism and clientelism and protect citizens from alien manipulation of consciousness while producing different levels of communication through endogenous progress. Nepali nations exist in many parts of the world and communicate with each other through language, culture and civilisation. This means that spatial view alone is not enough to arrest its historicity.

Now Nepali media are gripped by a dilemma between constitutional freedom and traditional politics rooted in tribal affinity, not national affinity with the state which is a sphere of rights, solidarity and democracy. Democracy has opened a scope for multi-channel message and citizens' feedbacks to leaders obliging them to evolve national outlook. The diversity of Nepali media has linked politics with social life - families, communities, cultural and linguistic groups, and articulated their desires so that illicit grievances do not upturn social cohesion.

Similarly, they have eased economic structures of production, exchange and distribution of public goods thus overcoming the problems of scarcity created by an imbalance of supply and demand. The current social dynamics in Nepal based on identity politics is, however, straining civic renewal facilitated by Nepal's many post-traditional societal traits vital to bond social capital for cultivating positive multipliers of nation-building. Some media in Nepal with distinct pattern of communication are more conducive to nation-building than the other.

Nepali media's role in building democratic life in society, political structures and parties and crafting a strong national identity cannot be misjudged but this type of media have yet to delink violence from politics. The control of the media by partisan, corporate and geopolitical interest tends to polarise the national audience into good and evil and chip the nation's civic culture.

Professional media grounded in public interests articulate distributive justice to foster a feeling of civic solidarity which creates general loyalty to constitutional state and adapt the old pattern of life to new stimuli and demands. They are considerate, not context free, in orientation as they help build thoughtful considerations among diverse communities, shape rational consensus and add to social cooperation by hauling positive message, information, values and issues that are essential for an educated public-policy oriented dialogue. In times of crisis they bring the connectors of society in a dialogue and synthesise diverse perspectives to fit into a national will. Nepal precisely requires a pro-active role of professional media which can foster a single public identity, equality of citizens before the law, instil civic duty to each other and close the elite-mass gap.

Enlightened media can pass civic knowledge in the entire arteries of polity, help each to bridge the space between the centre, Kathmandu and periphery and across the social classes and shore up the process of national unity. They help citizens free from fatalism, stamp out social ills of Nepali society and provide freedom from all types of tutelage. The sharing of lived experience of surviving citizens can provide learning opportunities for others to build on and strengthen resilient traits of national awareness, historical experience, crisis adaptation, revitalisation and post-conflict reconciliation.

Media freedom, in this sense, is connected to social accountability, public ethics and human rights. Likewise, pluralism in news means upholding normative standards of democracy, human rights, social justice and peace and the civic spirit of a deliberative civic culture which is vital to include the perspectives, values and interests of all sides in nation-building through diverse strategies and outreach:

The unification of people in the state promises its stability and endurance. In Nepal a gap persists between Nepali speaking, spiritually-oriented rural society and the English-stressing materially-oriented urban sub-society. This dualism, rested in a digital gap, is exposing the citizens to bumpy media socialisation and disposition without higher phase of life. Wearing away of this dualism is central to building this nation. Remote areas of Nepal demand the spur of progress indicators. Integration of the periphery to the centre, Kathmandu, requires the diffusion of media channels and effective outreach of the Nepali state so that democratic nation-building can impel from the bottom-up.

Nation building implies a process of cultivating common values, sentiments, feeling, emotion and interests across the diverse people. The split in education and health of Nepal, public and private, cuts up a critical support to under-classes and diffusion of cultural memories so crucial for political acculturation of diversity to changing times. Culture itself is a distilled knowledge shaped by communication. Nepali media must help to integrate political parties, civil society, business and citizens into a national communicative space and cultivate their loyalties to the state. So long as citizen, as members of state, is clientalised to personalised leaders building responsive capacity to state staggers. This is the reason Nepali media find difficulty to judge politics from non-partisan angle.

Self-governance is intrinsic to human genes. It is based on their self-developed consciousness about the physical space, language, culture, ancestry, history, religion and laws often reinforced by the narration of media. They together make up the nation of people. Modern nation-building requires the de-tribalisation of society and transfer of tribal, ethnic, caste, religious, regional and linguistic loyalty to the state, a .state having enough autonomy and capacity aided by its resources to determine its free existence. Tribalisation of political parties, media and civil society and their inability to imbibe civic virtues risk national cohesion in the same way as clientialisation of citizens to personalised leaders. Both cut the freedom of Nepalis and their ability to defend the hard-shell of the state within which democracy is situated.

The communicative gift of Nepalis is a key to realise their inner vigilance and ability to air voice on public matters. The free flow of information offered by the media under the Nepali constitutional right to information has made them critical of a range of issues and source of their valid judgment. Attentive citizens' expression in the social media reflect the condition of Nepalis and their needs offering opinions, options and promises for the solution of burning problems. Politicisation of citizens with added rights but few duties has, however, made Nepali politics aspiration-driven, even some media offer radical view rather than shape credible opinion and objective reporting essential for nation-building.

Media pluralism is the hallmark of democracy but diverse media socialisation of citizens has also fragmented Nepal's civic culture. When news, views and analysis are assorted with parochial interests, the flow of information turns public sphere jarring. Nepali citizens are, therefore, seeking to exonerate themselves from this sordid fact and criticise the national leaders for the nation's elongated political transition.

For nation-building Nepal has adhered the politics of social inclusion, partly multiculturalism, partly accommodation, federalism and promotional measures for weaker sections of society and quota for the marginalised. Nepal's distributive justice is rooted in a proportional electoral system in the structure of governance. The national construction of the nation demands a workable social contract whereby the state find easy to reconcile social fissures through peace dividends where Nepali media can assume communication, aggregation and articulation roles to link the state with citizens. Nepali media's role lies in moderating contesting views to a middle path and enables all people to use socialised choice.

The Nepali media have marked a shift from conventional watchdog role to a vibrant dialogue forum between the state and citizens. Only an interacting public sphere can draw citizens' policy attention. They can help to moderate the behaviour of legitimate, dissident and rebellious oppositions engaged in agitation and raise the voice of those exiled into to silence. Nepali media hold the power of public to improve the health of democracy and renew the society and the state to bounce back to a better condition capable of establishing a balance between public order, freedom, justice and peace.

Source: Dev Raj Dahal, Media as Nation Builder, \url{https://risingnepaldaily.com/opinion/media-as-nation-builder}

### Role of media duing crisis time

The COVID-19 pandemic has appeared as an unprecedented global crisis. Economic activities have shut down, social life has come to a standstill. More than the crisis, an innate psychological fear of the devastation that follows is deeply entrenched among public. With the World Health Organization already confirming that COVID-19 has affected 196 countries of the world, the gravity of the situation can be inferred. Various countries of the world have shut down with no signs of the pandemic coming to a complete halt. Although efforts are underway to find a vaccine for the cure of the virus, the real solution is looking unlikely in the near future. In this scenario, it is important that the state and citizens go hand-in-hand to minimize the adverse effect emanating from the COVID-19. The role of various state institutions including media and civil society will be pivotal to alleviate the crisis.
The manner in which the media is carrying the news and information pertaining to COVID-19 needs critical scrutiny. In our context, media seems to be creating deep confusion among the public about the real status of the coronavirus. While this is not to undermine the role of those media that are performing their duties responsibly, the issue is how to bring the entire media fraternity on track in term of abiding by the journalistic values. With an urgency to get themselves heard and seen over the rival media, some media have carried misleading headlines without verifying the source in the past few days. In the drive towards becoming first to give breaking news, some media have compromised the cardinal principles of journalism. One such case appeared this week when some media disseminated the news of a Birgunj resident being infected of the coronavirus which was later disproved by the concerned authority.
Another aspect of dissemination has to do with the choice of information and the manner of presentation. As the public is in a state of anxiety and fear during such difficult circumstances, media needs to provide content that increases optimism and hope among the public. This could in the form of making the headline news of the positive developments towards combating the coronavirus globally. Instead of reporting only the number of deaths as top news stories, successful stories of the virus treatment and the ongoing initiative towards finding a solution to this crisis can be the major headlines. Devising captivating promotional messages related to COVID-19 in a bid to generate mass awareness should continue.
Moreover, media can also carry videos of the successful treatment cases and the encouraging words of medical practitioners involved in curing the disease for public respite. Running interviews of psychotherapists and motivational speakers could alleviate the perturbed public mind. Broadcasting infotainment programmes related to history, litereature, music and sports could be interesting for the public.
Media can also support the state in its intiative towards combating the global crisis. Circulating the relentless efforts being made by the state will also serve the general welfare. Nonetheless, media as a watchdog of the state activities needs to bring into attention the public ethos. Bringing the stories of low wage laborers and other menial workers and calling for the state to address their hand-to-mouth problem could be one constructive way of media reporting during the crisis.
Since people are in self-quarantine at this hour of crisis with the government's decision of total lockdown at least for a week now, contents related to self-introspection, happiness and family relationships among others could appeal to the common people. With the predominnce of news and information pertaining to coronavirus, breaking the monotony also requires providing some refreshing materials to entertain the audience. In doing this, media ought to cater to the interests of segmented audience – children, youths and the aged.
As WHO has rightly pointed out, ‘infodemics'- an overabundance of information which exaggerates the severity of the problem and makes the solution difficult have posed a severe threat in the battle against the deadlly COVID-19. The need of a responsible journalism is all time high at the moment.

Source: Kushal Pokharel, \url{https://risingnepaldaily.com/opinion/role-of-the-media-during-crisis-time}

### Role of ICTs in Economic Growth

Information and Communication Technologies (ICTs), in the simplest way possible, can be defined as all the devices, applications and systems that allow people and organisations to interact in and with the digital world.

It is evidenced by case studies of various nations that ICTs can contribute to the national GDP and help improve market competitiveness of a country's goods and services. ICTs also have a positive impact on a country's governance and promote its international economic integration, improve living standards, narrow inequality and improve the utilisation of natural resources. All these functions of ICTs make it a key player in the 21st century in enhancing the economic growth of developing countries; because ICTs can help them make up for what they lack in industrial resources by catching up to the developed world in terms of information. This importance is also reflected in the fact that most of the countries today have national "e-strategies" and donor agencies routinely invest in them.

Now, in developing countries, there are mainly two ways in which ICTs contribute to economic growth. One is through their manufacturing and another is through their use. As ICTs are a combination of various individual technological entities, they require manufacturing. And most of this manufacturing is outsourced by the developed countries to the developing countries. This point can be illustrated by the examples of China and India. China is the manufacturing hub for Apple and India is the country which handles the software designing and packaging work outsourced by Microsoft. Such process of outsourcing helps provide decent jobs to thousands of people in the developing countries thus, contributing to poverty reduction.

Also, such companies not only bring their jobs but also their knowledge, which they impart on to the workers of the developing countries. This equips the labour force with skills which means that even if the outsourcing companies were to leave, the people would have necessary skills and expertise to get a job somewhere else or even start their own ventures. All of these serve to make the economy more productive and efficient – delivering economic growth.

Furthermore, setting up manufacturing points and industries in the developing countries means paying taxes to these countries which will directly contribute to the GDP and GNP. Also, we know that industries and especially ICT industries are not isolated entities but part of a larger value chain. Thus, the impacts mentioned above do not stay limited to one sector but branch to multiple sectors; all contributing simultaneously to the economic growth of a nation.

Then there is the act of buying and selling. ICTs, regardless of their place of manufacture, often always find their way into the markets of the developing countries where they are bought and sold. And just this simple act of exchange generates millions of dollars in revenue to the governments depending on the size of the market. This revenue is economic growth. This is how the physical aspects of ICTs help deliver economic growth in the developing countries.

But aside from these physical aspects, there is also the economic growth brought on by the use/application of ICTs. This can best be described through the following "five routes to economic growth:"

Economy: ICTs vastly increase the access to data and information crucial to the supply chain, at significantly reduced costs. Such increased data greatly increases the entrepreneurial activities of a country, thus increasing the efficiency and productivity. All of these end up promoting economic growth.

Efficiency: The use of ICTs can make an entire nation more efficient. The production sector becomes efficient as a labour force comprised of human activity complimented by technology becomes faster, longer and much more diligent than a solely human staff. Also, ICTs when adopted at the state level brings about "e-governance" which brings transparency, accountability and greatly streamlines the often-painstaking bureaucracy of the developing countries. This reduces corruption and makes entrepreneurial activities much simpler by removing the middlemen and the time-consuming hassles. Also, the need to make repeated lengthy journeys for activities such as meetings, ordering, delivering etc. is also reduced as ICTs bring about journey substitution. These, in turn, not only energise domestic commercial activity but also attract foreign investment. Thus, enlarging and enhancing the economy.

Profitability: When the cost of input decreases and the process becomes efficient then the output produced is of high quality. This output can then be sold at a high-value market (the information about which is also obtained from ICTs). Add to that, a transparent and corruption-free process, as discussed above, and this greatly increases the profitability of produced goods of a nation. This also helps in delivering economic growth.

Upgrading: Upgrading is changing the production process from a lower value to higher value ones. ICTs also help in this. With the use of ICTs, a business can take itself online where it may be able to expand its coverage area by getting more orders from different places. ICTs can also play a role in delivering these orders with technologies like drones being employed for the purpose. The payment process is also greatly modernised with various e-payment platforms being utilised. Thus, the whole process gets upgraded in value.

Restructuring: Restructuring means a shift of resources from a lower productive sector to a higher productive one. As discussed in the points above, ICT makes many sectors of the economy highly productive while at the same time, it opens up completely new sectors of production. Thus, investment of resources can be concentrated in these sectors which will yield maximum benefits for the economy of developing countries.

Thus, with the cost of inputs significantly decreased, production streamlined, outputs yielding higher profits, previously low-value sectors upgraded and entirely new sectors opened up for investments, this is how ICTs help deliver economic growth in developing countries.

However, this is not an unlimited scope and ICTs do have limitations, especially in the area of information failures. In ICTs, all routes to economic growth pass through the domain of information and unfortunately, this very domain of information is limited in developing countries by five types of information failures, namely absence, poor quality, uncertainty, asymmetry and cost of information.

Absence of information means that the key information that the development actors need such as the population of a place, literacy rate, sex ratio, livelihood data, internet penetration rate etc. are not available or at least the means to obtain such information are not available. Poor quality of information means that even if the information is available, they may not be of the required quality and usually not in an open format. Uncertainty of information suggests that the information may be available and may be of the desired quality but still, the accuracy of it may be questionable in developing countries; as most of the time the methodology will not be adequately given and the data collection and analysis procedure will not be clearly elaborated. Then, there is the problem of asymmetry where some development actors have access to key information that others lack. And due to the lax regulations and implementation in developing countries, such actors who do have the information start acting as middlemen and use that information for power and profit instead of passing it on to others and acting as facilitators for the economic growth.

And even if none of these information failures exist and quality and reliable information is available to all in an open format, even then, access to such information would be hindered by the cost. Many developing countries are still struggling to lift significant portions of their population out of poverty, the slightest bit of cost associated with information, such as the cost of travel, bureaucracy etc. will prevent people from accessing it.

Therefore, with the I (information) component of ICT facing such severe restraints, the role that ICTs can play in delivering economic growth in developing countries becomes limited.

But, all in all, it is of no doubt that ICTs have a significant role to play in delivering and accelerating economic growth. In the developing countries, this role becomes even more vital as they have tried and failed to do so with more traditional means. ICTs can be utilised in a two-pronged approach to bring growth to countries.

Firstly, the setting up of industries for their manufacture and the market activity generated by their consumption generate direct revenue to the government and also generate thousands of direct and indirect jobs.

Secondly, the use of ICTs helps facilitate start-ups and other entrepreneurial activities which have a broad impact on the whole value chain and contribute to poverty alleviation. ICTs reduce the costs of input and make the production process cheaper and more efficient. Thus, more value can be levied on the output which in turn makes the business profitable. Multiply these individual effects by thousands of businesses employing ICTs and then we get a macro-level national economic growth.

Source: \url{https://risingnepaldaily.com/detour/role-of-icts-in-economic-growth}

### Influence of communication in everyday lives

1. We live in a world besieged by media and technology. Every sector of our life is impacted by the presence of modern forms of communication system. Write an essay discusses the influence of modern forms of communications on everyday life.

\begin{solution}

The very nature of human being as a social creature underlines the importance of communication in one's day-to-day life. Human beings interact, possibly much effectively and efficiently than any other life forms in this planet. Much of the achievements we are relishing today, from kitchen utensils to high performance computers are all, in a way, fruits of reliable communication. This is to say that innovations are triggered in several ways through communication. It is also true that modern day communication, we use in the form of media and technology, have been made possible by the feats of human innovation -- once again guided by state of the art communication systems.

Early developments in communication technology consisted of invention of telegram and telephone, which extensively relied on electrical signals on copper wire. As the search for better communication infrastructure continued, data (including video and audio) sharing over wireless frequencies and optical cables are on upsurge today. To recapitulate the role of communication channel, we can refer to an event that is purported to have put an end to the World War II. Invention of cipher (a tool that decrypts hidden messages) on those days became a landmark in intercepting German radio communication channels and gaining the information about overhanging threat military assault. In the retrospect, communication development during World War might not seem as much of a feat when we consider high performing broadband facilites that are at our disposal today.

Communication system today pervades all spheres of life be it physical or socio-economic. It wasn't uncommon, just a few decades earlier, for people to commute great deal of distance just for the sake delivering information through postal services. Proliferation of telecommunication services in every major inhabitated part of world has made it non-essential for the paper mail services to physically transport tremendous amounts of exchanges people are making. This trend in frequent exchange of ideas and even general information among peoples -- families, friend, scholars, researchers, etc., is expected to only expand in future. The imperative is that, as means of communication gets cheaper and accessible, people need not be constrained to utilize information wealth due to their physical distance.

More often than not, means of communication exert their importance at times of crisis. Remote areas are remote for more than a reason but one of those consistently includes them not being able to communicate with authorities on time. This is a situation that typically results in greater marginalization, as individuals from distant localities lacking proper communication methods are increasingly unheard of their problems and scarcities. For example, when a disease outbreak occurs in such an area, relief package and medical interventions become delayed, resulting in significant losses of life. In contrast, well-connected areas where government can quickly initiate a rescue operation will have more lives saved.

When looked at for side-effects on an individual, communication affects even the mundane aspects of life. Citizens of 21 century take toolsets of communication for granted. There is little to how every day engagements of families are ran without crediting to modern communication technologies. Grocery stores and dairy shops have the communication network set to have knowledge of market; A dairy handler has to know who are his customers for what products. Daily updates on traffic and weather to work are all televised through various news channels so that commuters can be punctual to their work. Furthermore, in the workplace, all commercial startups communicate over telephone or email to their clients. This, in comparison to early days approach of door to door visit for sales promotion and order acquistion, is a stark gain in efficiency and scale. A supply order could be made over a phone call while also provisioning for accumulated order the previous day for the same household equipment to areas nearby thereby significantly cutting down the cost of repeated transportation.

Our social thread forms a web of close circle of friends and families along with tendencies to keep-on-touch with them. This emotional and social sphere of our lives too is strung more tightly together than ever before with the advent of new communication technologies. Rashid can now ask what the weather is like in Australia to one of his friends from Nepal in a video call, while also hoping some days to see them there in person. His mother can make a quick appointment with a doctor in a phone call after knowing that Birat is ill. In addition to surfing a personalized news-stand online right before the meeting, one is now also able to address a meeting which includes a mass of 50 people. This is made possible due to wide projection and affordable cinema technology which allow people to conduct presentations as well as entertain with movie.

Besides trivial occurances of everyday life, various channels of communication are used by government organizations to address public and state their vision for country. Likewise, several government insitutions have made social media formum also a platform for obtaining public feedback and outcries, while also attempting to address them. For example, if suspected for a criminal activity, goverment can call for a surveillance of an individual or a group through the use of closed circuit cameras (CCTV) or the GPS system of suspected's smartphone. Although for the case it is clear that criminal threat outweights the expectation that government action hurts individual privacy, there can be scenarios where this application of technology can be subjected to debate. Meanwhile, there is always a chance that political groups may misuse media to propagandize their statements to only gain popularity, while masking up facts that give different flavor.

With all the merits it has brought to individual and societal lives, modern communication technologies can well be thought of as integral part of development. While it has connected people for good, indiscriminate retrieval and exchange of some information can have detrimental consequences -- With great power, comes a greater responsibility.

\end{solution}
\end{questions}
